---
pdf_document:
  includes: null
  in_header: header.tex
date: "28/05/2021"
output:
  word_document: null
  pdf_document: default
html_document: null
title: "What are the effective features of consultation? A mixed methods approach"
always_allow_html: yes
bibliography: Thesis.bib
link-citations: yes
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#options(tinytex.verbose = TRUE)
#Install DiagrammeR to create the thematic map. It won't create a pdf without webshot and phantom.
#install.packages('DiagrammeR', repos = c("http://rstudio.org/_packages",
#                           "http://cran.rstudio.com"))
#library(DiagrammeR)
if(!"tinytex" %in% rownames(installed.packages())) install.packages("tinytex")
tinytex::install_tinytex()
#tinytex::tlmgr_install("pdfcrop")
#Sys.setenv(R_GSCMD="C:/Program Files/gs/gs9.54.0/bin/gswin64.exe")
#Will need the code below if I want to put in the diagrammer picture but don't currently need it.
#install.packages("webshot", repos = c("http://rstudio.org/_packages",
#                           "http://cran.rstudio.com"))
#webshot::install_phantomjs(version = "2.1.1",
#  baseURL = "https://github.com/wch/webshot/releases/download/v0.3.1/",
#  force = FALSE)
#will likely need the papaja for the write up in general (which has to be installed with the code below as it's not on CRAN yet)
devtools::install_github("crsh/papaja")
```

DO I NEED TO DEFINE THE WORD 'FEATURE'?

# 1 Introduction

This research consisted of interviews with Educational Psychologists
(EPs), the development of a novel questionnaire, and observations of
joint school-parent consultations with long-term follow-up. This was in
service of exploring what the core features of consultation are,
according to EP self-report and observation of real-world consultations,
and how EPs altered their practice to adapt to the COVID-19 global
pandemic. The interviews explored EPs definition of consultation, their
views on what the key features of effective consultations are, what some
of the barriers are, how they have changed their consultation practice
as a result of the pandemic and the advantages and disadvantages of
this. This was supplemented by a questionnaire which asked similar
questions, as well as asking participants to identify the different
kinds of work they were engaging in during the pandemic. The observation
schedule was informed by the relevant literature and was used to see how
often different features of consultation were observed during a joint
school-parent consultation and in what order. This was then to be
cross-referenced with reported progress towards jointly agreed goals for
the child and young people (CYP) to see which features correlate with
improved outcomes. This work built on a previous piece of research
exploring what EPs believed were the most important features of
consultation and a thematic analysis of recorded initial consultations
to identify the main features in a live consultation.

## 1.1 Literature Review

A literature review was conducted to see what previous research had
found to be the main features of consultation and what the main tools of
analysing the efficacy of consultation were. Various databases,
including Web of Science and Scopus, were searched using the key words
"educational psychology" and "consultation". Key references, such as
@kennedyEducationalPsychologistsWalk2008, were given to the researcher
by their supervisor to set a baseline for the literature review.
Considering all the relevant literature, there is some consistency
around EPs views regarding consultation. However, there is a
heterogeneity of understanding from other stakeholders as to what
consultation actually means. Crucially, there is a relatively small
amount of research exploring what happens during a consultation
[@kennedyEducationalPsychologistsWalk2008], as well as few studies
evaluating the efficacy of consultation. There are also few studies
which attempt to analyse what makes consultation effective or what the
effective features of consultation are. This leaves EPs and associated
stakeholders with a widely used but poorly understood and validated
framework.

### 1.1.1 What is consultation?

Consultation takes many different forms across contexts and countries.
Consequently, there is not a universal definition of consultation as
conducted by EPs. This raises an important problem for any EP who wishes
to engage in consultation or analyse its efficacy. Within a western
context, it fundamentally involves problem solving between consultants
(EPs) and consultees. The consultee is most often a teacher who knows
the CYP well, but it can also be parents and/or Special Educational
Needs Coordinator (SENCOs). In joint school-family consultations, it is
generally agreed prior to the consultation that at least one member from
the child's family unit and the school will attend. These individuals
collaborate to devise and establish interventions to help support and
find solutions for the client, the CYP
[@ofarrellResearchExploringParents2018]. Consultation is considered a
form of 'indirect' work as the theory is that the EP can enact the most
change for the CYP by meeting and working with those around the CYP
[@gutkinReconceptualizingSchoolPsychology1990]. They may or may not
engage one to one with the CYP but it is not mandated by this approach.

Consultation has become the model of service delivery for many
Educational Psychology Services [@sheridanRandomizedTrialExamining2017].
Most Educational Psychology Services (EPS) in the U.K. have moved
towards a predominantly consultation-based service
[@ofarrellResearchExploringParents2018]. This is in contrast with what
is viewed as a more traditional model which predominantly involves
individual casework, typically including the administration of a
cognitive assessment [@kratochwillEvidenceBasedInterventionsSchool2002;
@larneySchoolBasedConsultationUnited2003]. The most commonly employed
consultation framework in the U.K. is the Wagner model
[@wagnerConsultationApproachEducational1995;
@wagnerSchoolConsultationFrameworks1995;
@wagnerConsultationDevelopingComprehensive2000]. It is defined as "a
voluntary, collaborative, non-supervisory approach, established to aid
the functioning of a system and as inter-related systems"
[@wagnerConsultationDevelopingComprehensive2000] through "purposeful
[conversations] which [use] techniques of listening, clarifying,
problem-solving, challenging, questioning and reflecting"
[@munroAnglesDevelopingConsultation2000]. As a result, EPs work with
those closest to the CYP, but not as experts telling those directly
involved with the CYP how to help them. Their role is to help empower
the consultees to solve their own problems in school. The focus is not
only on the CYP but their relations with others and the many different
environments they are in, such as home, school, and their wider
community [@bronfenbrennerEcologyHumanDevelopment1981]. There is an
understanding of the interactions between these layers and the need to
consider a child holistically. This support is provided by asking
questions, analysing presenting problems and helping others think
differently, agreeing on potential interventions, and then reflecting on
the whole process so progress can be made.

### 1.1.2 How prevalent is consultation in the U.K.?

The move towards a consultation-based model of service is reflected in
government legislation. The Special Education Needs: Code of Practice
characterises consultation as one of the main services of EPs
[@departmentforeducationSENDCodePractice2015]. Several studies have also
found it makes up a large percentage of their time working with schools.
@shannonEducationalPsychologistEarly2007 delivered questionnaires to 44
EPs, asking for the EPs to self-report how often they undertook
different types of work, including consultation or case work. 32
responded, with most reporting they spent a majority of their time
engaging in individual level work. 91% of the EPs who were doing
individual level work stated consultation was the main activity
performed. However, the authors do not provide a definition of
consultation nor ask the EPs to provide a definition of consultation.
Given that consultation takes many different forms and there are a wide
range of views on consultation between EPs and other stakeholders,
ensuring everyone has the same definition of the process is crucial.
Without it, one cannot be sure different EPs are engaging in
consultation in a similar way and that the schools understand what they
are doing. Participants may have reported they used consultation, but in
practice their methods may be very different. Because of the limits of
self report, we do not know if such a disparity exists in this study. On
the other hand, the EPs who responded were from a large range of
locations across the U.K., thus increasing the representativeness of the
data.

Another study exploring the prevalence of consultation in the UK comes
from @leadbetterPatternsServiceDelivery2000 The authors sent
questionnaires to all Principal Educational Psychologists (PEPs) and
asked about their models of service delivery. Consultation was reported
as one of the most frequently used models. However, there was only a
return rate of 58%, with those not returning almost certainly not
randomly distributed. There is therefore uncertainty around the amount
of bias in the results. If the non-returns were randomly distributed on
key variables, such as whether the EPs has recently experienced a number
of organisational changes, then a low return rate would not introduce
bias into the results. But this is unlikely. As such, the results of
certain PEPs who may have different results from the norm are missing.

Although consultation forms the bedrock of many EPs work and the model
of service delivery for EPS, EPs often struggle to articulate what a
consultation model of service entails.
@leadbetterRoleMediatingArtefacts2004 argues there is little research
which explicates the structure and process of consultation. This
situation has not improved over the following years, with the few
studies examining this typically only focusing on one EPS or a small
number of EPs [@cordingStudyEducationalPsychologists2011;
@leadbetterInvestigatingConceptualisingNotion2006;
@pipherConsultationApproachTeacher2013a]. These limitations prohibit one
from developing a broad picture of how consultation is performed in the
U.K.

### 1.1.3 What are consultees views on consultation?

Before exploring what occurs in a consultation, understanding what those
involved (EPs, teachers) believe it to be is important. This is because
if consultees are to play an active role in consultations (as all models
of consultation state they should), their views of consultation need to
be understood. That way, any misunderstandings can be cleared up and
consultation can be effective. To serve this end, the following section
explore stakeholders' views of consultation as detailed in the relevant
scientific literature. @ofarrellResearchExploringParents2018 interviewed
three groups containing a teacher, an EP, and the parents of the child
the consultation was about. The teachers and parents reported that wile
they appreciated and saw the value of consultation, there was a lack of
understanding regarding its process and nature. All three teachers
implied they viewed the EP as the external expert, who had specialist
knowledge and access to resources which they wanted. This is in direct
contrast to the sentiments expressed by the EPs in this study. They
explicitly stated they were not experts and tried to distance themselves
from that sentiment. This concept is fundamental to many models of
consultation, including Wagner's. This research was conducted in the
Republic of Ireland. Here, consultation has only become the dominant
model of service delivery in recent years. Thus, U.K. based teachers and
SENCOs may have a better understanding. However, few pieces of research
have been conducted to explore understanding of this important strand of
EP work.

@dennisFarGoodQualitative2004 interviewed SENCOs at twelve schools to
better understand their views on EP work. One of the core themes raised
by the participants was a wish for EPS to "publicise more the range of
things it has to offer and good/innovative practice". The exact number
of schools which held this belief is unknown as that information was not
reported in the paper. Regarding consultation, there was a large
heterogeneity in the the school's understanding of it. Some schools were
fully aware of the consultation model of service delivery and preferred
it to previous models. Such models focused on direct work, including
assessment of CYP using cognitive assessments. However, other schools,
either did not understand the consultation model or were only partially
aware of it but preferred other models which placed a primacy on
child-focused individual work. They reported they believed the
consultation model involved "too much talk, not enough action"
[@dennisFarGoodQualitative2004, 22] and thus did not value it as highly.

This lack of understanding of consultation is found in other countries
as well. Many Australian EPS have also shifted their focus from an
assessment-based to a consultation-based model of service delivery. But
they also experience a lack of cohesion in understanding among
stakeholders regarding the meaning and process of consultation
[@bellPerceptionsRealitiesRole2013]. However, some EPs do not report
this problem. In the U.S.A., those who work with school psychologists
(as EPs are called) show a greater consistency of understanding of
consultation in schools. There is also a larger evidence base for the
efficacy of their form of consultation. This is because consultation as
practised in the U.S.A. is almost exclusively Conjoint Behavioural
Consultation (CBC). CBC is defined as "a strength-based, cross-system
problem-solving and decision-making model wherein parents, teachers, and
other caregivers or service providers work as partners and share
responsibility for promoting positive and consistent outcomes related to
a child's academic, behavioural, and social--emotional development"
[@sheridanConjointBehavioralConsultation2007, p.25]. CBC has been shown
to be effective for CYP over a wide range of settings and for many
presenting problems [@sheridanRandomizedTrialExamining2017]. This
hegemony of practice allows for a consistent definition and
implementation, and is likely one of the reasons consultation in the
U.S.A. is better understood and valued by stakeholders
[@reddyClinicalFocusConsultation2000]. It also means its efficacy can be
assessed more readily, such as by conducting a randomised control trial
conducted by @sheridanRandomizedTrialExamining2012

However, the lack of understanding by key stakeholders (SENCOs,
teachers, and parents) may not truly reflect the modern day conception
of consultation in the U.K. The research reporting these findings are
roughly 15 years old. It is reasonable to presume stakeholders have
become more familiar with it, given how prevalent it is. A more recent
paper exploring this is @cordingStudyEducationalPsychologists2011. For
this work, the authors interviewed 10 school personnel (such as Head
teachers and teachers) and 9 EPs in a Welsh Local Authority (LA). The
aim was to elicit their understanding of the kinds of work they believed
EPs engage in. There was a general alignment between the views of the
school personnel and the EPs themselves. But the school personnel stated
they greatly valued the EPs expertise in diagnosing and alleviating
presenting problems. This shows that despite there being a shared
understanding of what EPs do, there is still a divide in what
stakeholders value about EP work.

### 1.1.4 What are consultant's views on consultation?

The vast majority of EPs have a positive view of consultation, with the
Local Offer literature from many EPS stating their model of service
delivery is consultation, such as Kensington & Chelsea
[@royalboroughofkensingtonchelseaEducationalPsychologyService2019]. Some
EPs believe they provide a unique method of working through their use of
consultation [@ashtonWhatValuableUnique2006].
@ashtonWhatValuableUnique2006 sent questionnaires to both schools and
EPs asking for them to comment on the work EPs engage in. 22
questionnaires (out of 58) were returned from mainstream primary schools
and eight (out of nine) EPs completed the questionnaire. 'Statutory
assessment work', 'Closed tests' and 'Expert role' were the most
frequently provided parts of EP work that were classed as unique by
SENCOs. Few or no other agencies were judged by SENCOs as providing a
similar or the same service as EPs. 'Individual assessment and
intervention' and 'Consultation' were the aspects of EP work the EPs
themselves believed were unique to them, as no other service provided
these. This shows a clear disparity between the views of EPs and key
stakeholders within consultation (SENCOs). They also reported that
SENCOs typically valued more traditional EP work, such as individual
assessment and giving advice. The SENCOs did not value consultation, nor
give evidence they had a complete understanding of it. However, these
results should be interpreted with caution. The small sample limited to
one LA undermines our ability to generalise the results to a wider
context. It also only collected data from mainstream primary schools,
further limiting the scope of these results. Yet one of the main
results, namely the lack of understanding regarding the nature of
consultation) has been replicated by research in other school settings
[@dennisFarGoodQualitative2004; @ofarrellResearchExploringParents2018].

### 1.1.5 What are the main features of consultation?

Once a common understanding of what the stakeholders believe
consultation to be has been created, an analysis of the common features
of consultation can occur. The following section will evaluate the
relevant literature regarding the features of consultation.
@hendersonExplorationImpactConsultation2013 used focus groups with
clusters of SENCOs across a small LA to gain an understanding of their
beliefs about the mechanics of consultation. The researchers sat in on
five different Primary SENCO Network meetings and worked to elucidate
their views. They presented the participants with statements about parts
of the consultation process. Their task was to sort them depending on
how often they believed the statements to be a part of a consultation.
The mostly commonly given features of consultation were: discussing
issues with relevant parties; information gathering; and it being a
reflexive process with a focus on collaboratively crafting solutions.
They also conducted semi-structured interviews with EPs, children who
were receiving EP involvement, and their parents. It being a
collaborative and problem-solving process, with a focus on solutions,
and the development of positive working relationships between those
involved were the two main themes. The use of focus groups to identify
SENCO beliefs regarding the nature of consultation and the interviews
with the stakeholders allows comparison between the stakeholders
expectations and the reality of consultation. However, given the
researchers did not directly observe consultations but relied on
self-report, the conclusions that can be drawn regarding how
consultations occur are weakened. This is because of the disparity
between self-reports of behaviour and real-world instances of behaviour
[@argyrisTheoryPracticeIncreasing1992].

@kennedyEducationalPsychologistsWalk2008 thematically analysed the voice
recordings of 17 EP-teacher consultations. These individual case studies
[@robsonRealWorldResearch2015] were supplemented by a pre-consultation
questionnaire completed by EPs to establish their espoused theory for
consultation. A comparison could then be made between the recordings and
participants' self-report to see whether their espoused theory aligned
with the recorded behaviours. The authors report a high correspondence
between the EPs espoused theory and theory in practice as EPs
predominantly engaged in behaviours dictated by their espoused theory.
The most common behaviours by EPs were working collaboratively,
typically with those most involved (predominantly parents) using either
Solution-Focused approaches or problem-solving analysis.
Solution-Focused approaches are characterised by greater interest in the
solutions to presenting problems rather than the problem itself. It
views the client as capable of solving their own problems with a changed
mindset, facilitated by the EP, through identifying times when the
severity of the problem is reduced or it is not present, termed
'exceptions' [@rhodesSolutionFocusedThinking2004a]. Problem-solving
analysis is related to behavioural consultation
[@berganBehavioralConsultationTherapy1990] and is divided into four
stages: problem identification, problem analysis, treatment
implementation, and treatment evaluation
[@sheridanSchoolConsultation2000]. Those involved with the young person,
such as teachers, are involved throughout
[@kennedyEffectiveConsultationEducational2009a]. By recording the
consultations, the authors could gather data from a larger number of
consultations than they could have if they sat in on every consultation.
However, there was a low granularity of analysis. The researchers only
assessed whether features of the espoused theory appeared at all during
the consultation. Thus, the analysis only shows that during a
consultation, EPs brought in ideas from their espoused theory at least
once. There was no analysis of how frequently the ideas appeared and
when during the consultations. It therefore cannot tell us how great a
part these concepts from the espoused theory played in the consultation,
merely that they were present.

@nolanProcessPsychologicalConsultation2014 observed seven consultations
between five EPs, a teacher, and at least one parent. A week later, the
researchers conducted semi-structured interviews with all EPs and
teachers and some of the parents. Several key themes arose from the
observations and interviews. These were: empowering those involved in
the consultation; working collaboratively; the importance of each
participant in the consultation recognising the valuable knowledge from
others; reviewing outcomes; and EPs using their expertise to support
others (without emphasising their role as the "expert"). The use of both
observation and interview generates a lot of data about these 7
consultations, giving a very detailed understanding of the process. It
also allows corroboration between data collection methods. However, the
small sample size limits the generalisability of the findings. These
results therefore need to be replicated with different configurations of
participants and in other school settings.

@ofarrellResearchExploringParents2018 found teachers appreciated
consultation as they felt empowered to support the pupils who had been
referred. According to @jonesRefocusingEducationPsychology1990, this
empowering of consultees rather than fixing the consultees problems or
simply giving advice, is part of the definition of consultation.
@dennisFarGoodQualitative2004 found that EPs and SENCOs saw several key
issues relating to the successful implementation of a consultation
model: the EP having a detailed knowledge of the system (school) they
were working in; addressing issues at multiple levels (rather than just
on the individual level); positive relationships between the EP and
SENCO; and empowering staff to successfully fix their problems, rather
than doing it for them.

@dickinsonConsultationAssuringQuality2000a &
@munroAnglesDevelopingConsultation2000 examined how consultation had
been implemented in their EPS (Lincolnshire and Buckinghamshire
respectively). Behaviours and approaches which helped support the
successful implementation of consultation across both EPS included:
having purposeful conversations; EPs using their psychological knowledge
during consultations; and all parties involved deciding on interventions
as well as reviewing past and current interventions. Factors that were
only reported in @munroAnglesDevelopingConsultation2000 included:
engaging in preventative work; improving outcomes; and engaging in
multi-level collaborative work. Unfortunately, these papers are based on
the author's reflections on the implementation of consultation in their
LA and do not give the views of anyone else or provide much in the way
of data to support their findings. Readers must therefore take them at
their word.

This research builds on a previous piece of work by the lead researcher.
This first work explored what EPs believe the key features of a
consultation are and what happened in an initial consultation between at
least an EP and a school staff member. This was done through a novel
questionnaire asking EPs to rank features of consultation according to
their importance and thematically analysing transcripts of
consultations. During the consultations, the two most frequent features
of consultation were 'Understanding the presenting problem' and 'Working
together to come up with solutions'. EPs rated these as core features of
consultation in the questionnaire, as well as improving outcomes for
young people. Whilst this research assessed what EPs believe the core
features are and what the features are of an initial consultation, the
small sample size (3 observed consultations and 8 EPs completing the
questionnaire) means the results are hard to generalise beyond the
immediate consultations.

Although these studies typically only focused on a small number of
participants, the consistency in results allows fundamental features of
consultation to be gleaned. The studies also cover a wide range of EPS,
so the results are not limited to a specific region. This increases the
generalisability of the findings. However, despite these consistencies,
there is still a great deal of heterogeneity in consultation models and
practice. EPs can state they are engaging in consultation, but without
more information or a previously established working relationship, those
involved (parents, teachers, etc.) are unlikely to know what to expect
with a consultation. An arguably more serious consequence is that
assessing the efficacy of consultation is very difficult. If
consultations are not ergodic due to the very wide range of features,
any assessment of consultation may not be valid for consultations
performed by an individual EP. Therefore, assessing the efficacy of
consultations is difficult. This is against the backdrop of EPs working
within 'traded services' [@leeExplorationDevelopingRole2017], where the
ability to demonstrate efficacy is highly valued. It therefore behoves
EPs to gain an understanding of the consistent features of consultation.
This will allow some assessment of which features are correlated with
improved outcomes for CYP.

### 1.1.6 Assessing the efficacy of consultation

There have been calls for assessing the efficacy of EP work for decades,
such as @clineQualityAssuranceEducational1994a, but this has become even
more important since the almost complete shift to 'traded services'
[@nationalcollegeforteachingandleadershipEducationalPsychologyWorkforce2014].
'Traded services' marks a shift in funding towards existing service
organisations needing to generate income from schools (seen as
customers) to either partially or fully financially support itself
[@woodsPreparationPractitionerEducational2014]. Many EPs feel a pressure
from schools to both provide something tangible for customers and to
demonstrate the effectiveness of their work, so schools buy their
services again [@leeExplorationDevelopingRole2017]. EPs are also
expected to use evidence-based tools and to critically evaluate their
practice [@britishpsychologicalsocietyQualityStandardsEducational2015,
Standard 4.8;
@healthcareprofessionscouncilStandardsProficiencyPractitioner2015a,
Standard 12.1] as part of the requirements of practising as an EP. It is
therefore very important for EPs to understand what aspects of
consultation are effective in eliciting change.

Measuring such change is difficult. As
@kennedyEducationalPsychologistsWalk2008 notes, due to the complex
nature of the interactions between consultant, consultee, and client it
is difficult to decide what to measure and how to do so. Several methods
have been put forward but none have gained ascendancy yet. One method
used by some EPS [@hampshireepsHowEducationalPsychology2010] is the
Target Monitoring Evaluation [@dunsmuirEvidenceBasedPractice2009].
Target Monitoring Evaluation (TME) is based on Goal Attainment Scaling
(GAS), which was developed by @kiresukGoalAttainmentScaling1968 to
evaluate the outcomes of mental health interventions. TME is a
streamlined form of GAS, with the added advantage of increased
granularity in evaluating progress in relation to expected progress. TME
involves the negotiated development of SMART goals (specific,
measurable, achievable, realistic, and time limited) between the EP and
the consultees. TME forms were completed by both EPs at two Local
Authorities and assistant EPs in one County. During the initial
consultation, after the goals had been decided upon, each participant
rated how far along on a 10-point scale the child currently was towards
each goal. They then stated how far they expected the child to be when
they had their review consultation. 6-8 weeks later, during the review
consultation, each participant rated how far the child had actually
progressed, which was compared with how far they were predicted to
progress. Interviews were conducted with teachers, SENCOs, and
headteachers, who gave positive feedback on the easy and efficiency of
the process, as well as how the tool helped focus on setting of targets.
Two of those interviewed had experience with GAS and stated they
preferred TME. Focus groups of EPs and assistant EPs elicited positive
views towards the tool, as well as considerations of implementation.

This pilot study gives evidence for TMEs efficacy in assessing progress
in response to EP intervention. However, the limited detail provided in
the report means we do not have a fine grained understanding of the
strengths and weaknesses of the tool. @monsenEvaluationPreTraining2009
assessed the efficacy of assistant EPs using TME and focus groups with
stakeholders. Both measures found assistant EPs to be beneficial to EP
work. This work was conducted in one EPS, therefore the generalisability
of the findings are limited. TME is a quantitative assessment of
efficacy and the focus groups produced qualitative data, comparison of
efficacy between the two measures is difficult. We therefore cannot draw
strong conclusions about the validity of TME when assessing educational
psychology work.

There have been a few studies which have attempted to compare TME with
other quantitative measures of change, such as
@connorTargetMonitoringEvaluation2010. In this thesis, the author
compared TME with other, more established forms of progress measurement
in domains like reading, such as the York Assessment of Reading
Comprehension (YARC). They report that there was broad agreement between
the TME and other forms of assessment; when other forms of assessment
found improvement, this was reflected in the reported change through the
TME forms. However, while TME may be useful for identifying progress in
individual children, it was not clear how it could be used to assess the
quality of the work from the EP. There were also some difficulties
regarding the use of it, as there was disagreement between some
consultants and consultees regarding target setting and the voice of the
child.

A recent paper [@eddlestonUsingProfessionalPractice2018] comparing
different consultation evaluation tools excluded TME because it did not
reach the inclusion criteria. Here, TMEs streamlined nature counted
against it as it was not sufficiently thorough enough to be evaluated.
This means there is limited evidence for its efficacy as a tool.
However, as @dunsmuirEvidenceBasedPractice2009 states, "the strengths of
GAS are maintained but the TME system is more streamlined and user
friendly" (p 67). We can therefore have increased confidence in the
validity of TME as a measure of change, given that GAS has been shown to
be a useful tool [@roachGoalAttainmentScaling2005a] and it shares
fundamental similarities.

### 1.1.7 Local Offer literature

To gain an understanding of what EPs at different LAs understood
consultation to be, the Local Offer literature was examined. This
information was found on the LA's websites and detailed what services
the EPS provided. Despite almost all services having moved to a
consultation-based service delivery
[@dinkmeyerConsultationCreatingSchoolBased2016], over a third of LAs did
not explicitly mention consultation. Of those that did, the most
commonly cited feature was working with relevant parties, such as
teachers. The second most common was improving outcomes for the CYP,
with the importance of looking for solutions (including the use of
Solution-Focused approaches) also being mentioned frequently. What this
shows is that for the LAs that mention it, the EPs working there have
explicitly stated the importance of collaborating with those closest to
the CYP and the necessity of improving the CYP's outcomes.

## 1.2 Context and rationale

This project was greatly shaped by the coronavirus (COVID-19) pandemic
and the subsequent response by the British Government. Because of this,
the research was conducted during unprecedented circumstances. All EPs
(and workers in general) had to work remotely from home. This presented
unique concerns for how EPs worked, as they were not allowed to see any
adults or CYP in person. Several documents, such as
@bhardwajRapidLiteratureReview2020, and one piece of research
[@aepSurveyEffectsCovid192020] were disseminated drain this time,
detailing guidance as to how EPs can work ethically within the context
of a lockdown. This included conducting consultations using either
phones or video call software, such as Microsoft Teams or Zoom. There
were concerns regarding the safety and privacy of Zoom technology
[@paulZoomMalwareWhy2020] so Teams was encouraged by many EPS. This
shifted the manner in which the research could be conducted:
consultations could not be observed in person and many EPs were not
engaging in consultation. The timeline of the research was changed as a
result, with the observation of consultations pushed back to September
2020 when it was hoped they would resume by. There was also a broadening
of the parameters of the research. Questions exploring the impact of the
lockdown were posed to interviewees and a questionnaire was designed and
disseminated to explore a wide range of views towards consultation and
how it had been affected by the pandemic.

The purpose of this research was it gain an insight into what happens
during a joint school-family consultation, as well as which features
correlate with rated changes towards agreed goals. Having a more
fine-grained understanding of when different feature are seen and how
frequently would provide valuable insight into what are the core feature
of a consultation. This could then be cross-referenced with the ratings
of progress as measured by TME. Because of the global pandemic, a
related question was explored regarding the use of technology when
conducting consultations. Gaining an understanding of what the core
features of consultation is allows EPs to understand what is essential
for a consultation to occur. This will inform the professions response
to the lockdown, subsequent lockdowns, and EP work in general. EP
beliefs regarding how consultation can be performed and their
experiences changing their work can give guidance as to how EPs should
use consultation in future, beyond the scope of the COVID-19 pandemic or
any future pandemics.

## 1.3 Research questions

Given the lack of strong theory in this area of research, research
questions were developed but statistical hypotheses could not be drawn.
Thus, it is exploratory research
[@kimmelmanDistinguishingExploratoryConfirmatory2014]. The research
questions are:

1.  What do EPs believe are the core features of a joint school-parent
    consultation?

2.  Which features of consultation correlate with increased progress
    towards agreed goals?

# 2 Methodology

## 2.1 Epistemology and research paradigm

To explore these questions, a mixed methodology was employed, making use
of quantitative and qualitative research methods. It was informed by a
scientific realism epistemology. Scientific realism can help approach
difficult problems in social science as it takes into account the
complexity of the situation in which they occur
[@houseRealismResearch1991]. It can be viewed as a pragmatic approach
[@robsonRealWorldResearch2015] as it is less concerned with
philosophical dualisms, such as rationalism versus empiricism, and more
with practical considerations of issues and potential solutions
[@johnsonMixedMethodsResearch2004a]. Mixed methodology aligns with a
pragmatic approach as it is not beholden to one method of exploring a
research topic; it sees the benefits of both for exploring a research
question in different ways [@denscombeCommunitiesPracticeResearch2008].
Multiple methods of inquiry were employed because it is generally
believed using different means to explore research questions brings
greater rigour [@creswellResearchDesignQualitative2003a]. Data can be
triangulated with one another, with evidence corroborating, refuting, or
adding nuance to each other and increasing confidence in ones findings
[@munafoRobustResearchNeeds2018]. Mixed methodology research designs can
be divided along a key dimension: paradigm emphasis
[@johnsonMixedMethodsResearch2004a]. This refers to whether one strand
of the research (quantitative or qualitative) is given greater emphasis
during analysis. Because equal weight was placed on both forms of
inquiry, this was an 'equal weight' paradigm emphasis piece. An explicit
account of the ways in which the qualitative and quantitative arms of
the research relate to one another will be given
[@denscombeCommunitiesPracticeResearch2008].

## 2.2 Participants

Ethical approval was obtained from UCL the Institute of Education's
Ethical Committee. The inclusion criteria for all three arms of the
research was: an EP or TEP who used consultation as part of their
practice. There were no requirements as to how frequently or recently it
had to be used, nor experience or location. Nor were there requirements
around the definition of consultation; just that EPs believed themselves
to be engaging in consultation. This was to try and elicit a wide a
range of views on consultations from practising EPs. For the interview
and observation, participants were recruited via the researcher's EPS.
Convenience sampling was therefore used. This was because participant
recruitment for the observation was judged to be difficult and the
researcher would have greater success by asking participants they
already had a professional relationship with. The interview also
recruited participants by sharing a call for participation on a popular
mailing list for EPs and other education professionals (EPNET) and
social media (Twitter). Participants were also asked to share the call
for participants with other EPs at their work. This was the method used
for recruiting participants to the questionnaire as well. Thus, a
mixture of convenience and snowball sampling
[@robsonRealWorldResearch2015] was employed for these two arms of the
research.

## 2.3 Materials

All materials, along with raw data, are released under a CC-BY license,
thus allowing re-use of materials and improving reproducibility and
transparency [@nosekScientificUtopiaII2012a]. They can be accessed at:
<https://osf.io/nra86/> in the 'Methods' folder. Almost all materials
used were Free/Libre and Open Source Software [@stallmanFLOSSFOSS2016].

### 2.3.1 Interviews

A semi-structured interview format was used because an interview
schedule was developed (Appendix 1) which served as a checklist of areas
to be explored with a given question order and wording. However, the
order and wording was allowed to change given the flow of the interview.
Additional questions were used to further develop a interviewee's answer
[@robsonRealWorldResearch2015]. The interviews were of the focused type
as the questions centred around the key theme of consultation
[@mertonFocusedInterviewManual1990]. Probes (interview devices to elicit
more information) were employed by the researcher to further develop the
interviewee's responses. To achieve this, 'laddering questions'
(questions phrased in a variety of ways asking for the interviewee to
expand on their answer) and 'summarising techniques' (summarising what
has just been said by the interviewee to prompt more information), as
well as 'addition probes' to maintain the flow of the conversation
[@zeiselInquiryDesignEnvironment2006]. All interviews were recorded with
an Honor 10 lite phone. The anonymous transcript was thematically
analysed using the software NVivo .

### 2.3.2 Observation

The quantitative arm of the research involved systematic observations of
joint home-school consultations with an EP. Thus, it was a naturalistic
observation as the participants were observed in their typical
environment without any interference from the researcher
[@viglioccoTipofthetonguePsychology2001]. A systematic observation was
chosen as it helps overcome the often recorded discrepancy between what
people say they do and how they behave in the real-world. This has been
reported in such wide-ranging fields as smartphone use
[@andrewsSelfReportToolsCompare2015a] to driving behaviours
[@kayeComparisonSelfreportObjective2018]. They involve the development
of a coding scheme (Appendix 2) to identify categories over the course
of a set period of time. The categories are defined and operationalised
prior to data collection [@crollSystematicClassroomObservation1986].
They were derived from the relevant literature and were mutually
exclusive. The categories were limited to what was explicitly said.
Models of consultation, such as Solution-focused and problem-analysis,
were broken down into their constituent observable parts, such as
exploring strengths and identifying exceptions. Event sampling was used
as the absolute and relative frequency of events was of interest
[@robsonRealWorldResearch2015]. A sequence record was also used to
provide information as to the order in which the features were seen,
thus providing information about transitions
[@robsonRealWorldResearch2015]. Time sampling was not chosen so no
events were missed because they fell outside of the time intervals.
However, the length of time each feature occurred for was lost. Whilst
this information would be valuable to see how long each feature lasted
for, rather than just how frequently it occurred, it was decided that
the risk of missing feature due to the researcher focusing on correctly
marking the time of each feature outweighed the benefit of gaining that
information.

To explore the relationship between the features and the change in the
goals for each consultation, Qualitative Comparison Analysis (QCA) was
used. It is characterised as a "small-N-many variables" approach.
Configurational Comparative Methods: Qualitative Comparative Analysis
(QCA) and Related Techniques: "QCA techniques allow the systematic
comparison of cases, with the help of formal tools and with a specific
conception of cases."

"In the process of configurational comparative analysis, the researcher
engages in a dialogue between cases and relevant theories. Indeed, the
choice of the variables (conditions and outcome) for the analysis must
be theoretically informed. In this sense, there is a deductive aspect to
QCA; however, QCA techniques can also be used more inductively, gaining
insights from case knowledge in order to identify the key "ingredients"
to be considered (Rihoux, 2003, 2006; Rihoux & Lobe, 2009)."

@rihouxCaseQualitativeComparative2009

"QCA techniques allow for "conjunctural causation" across observed
cases. This means that different constellations of factors may lead to
the same result (equifinality)"

"By using QCA, the researcher is urged not to specify a single causal
model that best fits the data, as one usually does with statistical
techniques, but instead to determine the number and character of the
different causal models that exist among comparable cases (Ragin,
1987)."

for the statistical programming language R
[@rcoreteamLanguageEnvironmentStatistical2017]

<https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1089061/pdf/hsresearch00022-0148.pdf>:
"to assess the sufficiency of a combination of causal conditions, the
researcher selects cases with a given combination of conditions and then
evaluates whether or not these cases display the same, or roughly the
same, outcome."

### 2.3.3 Questionnaire

A questionnaire was designed using Qualtrics to explore consultation as
conducted both during the lockdown and prior to it. The questions were
informed by the answers to the interview questions, as recommended by
[@gehlbachMeasureTwiceCut2011]. Due to the fact the way answer options
are presented can bias results [@schwarzSelfreportsHowQuestions1999],
the construction of the question and answers was guided by the best
practice recommendations from @gehlbachMeasureTwiceCut2011. Quantitative
questions explored what the key features of consultation are (both
before and during the lockdown), what kinds of work EPs engaged in
(before and during the lockdown), how they have found the changes to
their work (with Likert scale ratings), and how much different types of
work have been affected by the lockdown. Questions suggesting different
features of consultation were based on the scientific literature, for
example @dennisFarGoodQualitative2004;
@dickinsonConsultationAssuringQuality2000a;
@farrellDevelopingInclusivePractices2006;
@hendersonExplorationImpactConsultation2013;
@kennedyEducationalPsychologistsWalk2008;
@munroAnglesDevelopingConsultation2000; &
@nolanProcessPsychologicalConsultation2014, and the material on the
Local Offer websites. These were the same features used in the
observation schedule. Open-ended questions will be used to explore the
changes to EP consultations as a result of the lockdown and their views
towards technologically mediated consultations. This is because there is
no published literature to suggest what EPs may experience, given the
unprecedented nature of the present circumstances. It is therefore best
to give participants the opportunity to respond how they wish, without a
narrowing of options by the questionnaire.

## 2.4 Procedure

Prior to data collection, it was decided the quantitative arm would be
conducted first, starting in March 2020 and continuing until March 2021.
Interviews would be conducted in the autumn of 2020. Thus, a concurrent
triangulation design would be employed
[@creswellResearchDesignQualitative2003a]. Both the quantitative and
qualitative arms of the research would be conducted simultaneously and
independently. The results were to be compared to see whether the
conclusions drawn align with one another. This was done for practical
rather than philosophically informed reasons. It was agreed beforehand
that collecting observation data would be more difficult, as finding
consultations with all the required participants who were also willing
to be observed is unlikely. Consultations with both the teacher and
parent present are less likely to occur than consultations with just one
of them, given the specific multiplication rule of probability
[@grinsteadIntroductionProbability1997]. It was therefore felt that
having a longer window of opportunity to collect data was the reasonable
course of action.

However, due to the COVID-19 pandemic all in person consultations were
cancelled across the U.K. to comply with the government-mandated
lockdown [@cabinetofficeStayingHomeAway2020]. Whilst many EPs offered
consultations to their respective schools, most found they delivered far
fewer consultations during the lockdown than usual. Those who delivered
consultations typically did so via the phone, eliminating any chance of
observation by the researcher. In response, data collection for the
interviews was brought forward to start in March 2020 and observations
of consultations would occur once consultations could be observed by the
researcher. The research was therefore adapted to use a sequential
transformative design [@creswellResearchDesignQualitative2003a]. This
type of mixed methodology involves one method preceding the other.
Either the qualitative or the quantitative arm of the research project
is conducted first. The methodology does not require one be used before
the other, so practical reasons may determine the order of research. The
results from both strands are interpreted together, with one informing
the other.

### 2.4.1 Interviews

Interviews were originally planned to be in person with EPs in the
researcher's EPS. However, because of the global pandemic, all
non-essential in person meetings were banned. They were therefore
switched to video or phone call interviews. Because of the sudden
increase in proficiency and willingness of many EPs to use phone and
video call technology, the parameters of the participant recruitment for
the interviews was widened to all EPs. This decision was made because of
a desire to increase the number of participants and thus the range of
views on consultation.

Semi-structured focused interviews were used to elicit EP views with
regards to the core features of consultation, the barriers to effective
consultation, how their consultation work has changed in response to the
lockdown, and the advantages and disadvantages of this new way of
working. 30 EPs were interviewed using a mixture of phone and video call
technology. Data collection took place between 31/03/2020 and
28/05/2020. All interviews were recorded and an anonymous transcript
made. These transcripts were thematically analysed, which involves the
identification of themes through "careful reading and re-reading of the
data" [@riceQualitativeResearchMethods1999]. A mixed or hybrid thematic
analysis approach [@feredayDemonstratingRigorUsing2006] was employed.
This incorporates inductive and deductive thematic analysis. Inductive
thematic analysis is driven primarily by the data
[@boyatzisTransformingQualitativeInformation1998a] and deductive
thematic analysis is theory-driven with codes derived from said theory
[@crabtreeTemplateApproachText1992]. The a priori codes identified were
developed from the scientific and Local Offer literature. Semantic
themes (that which is explicitly said) were found and analysed
[@boyatzisTransformingQualitativeInformation1998a].

### 2.4.2 Observation

After gaining informed consent from all participants, the researcher
observed the consultation unfold as normal. The researcher will use the
observation schedule to mark when and how frequently different features
occur. These will then be summed. Immediately after the conclusion of
the consultation, each participant (EP, school staff member, and
parent/guardian) was asked to collectively identify 2-3 goals for the
CYP to work towards. This was done using a TME form. Participants rated,
on a scale of 1-10, where the CYP currently was towards that goal (by
writing the letter 'B' for 'baseline' next to the number) and where they
expected them to be in 6-8 weeks (by writing the letter 'E' by the
number). In 6-8 weeks time, participants would be contacted by the
researcher via email to rate how far along the CYP had progressed
towards that goal. This judgement was represented by the letter 'A' (for
'actual') along the same rating scale. This data will be summarised with
the median ratings for each category ('baseline', 'expected', and
'actual') presented. The summed features would be tallied against the
progress made for each observed consultation. This will be calculated by
subtracting the 'baseline' rank from the 'actual' as research suggests
most TME forms report a positive change as a result of the consultation
[@dunsmuirEvidenceBasedPractice2009; @monsenEvaluationPreTraining2009].
Whilst the independent variable of frequency counts of features is
continuous, the dependent variable of the reported progress is ordinal.
Therefore, a Spearman rank-order correlation will be used
[@fieldDiscoveringStatisticsUsing2012] to measure the correlation
between features of consultation and reported change in outcomes for the
CYP.

### 2.4.3 Questionnaire

A questionnaire was designed to supplement the findings from the
interviews. By identifying key ideas from the interview answers, these
could be explored with a larger sample by using a questionnaire. Data
collection took place between 26/05/2020 and 04/06/2020. Descriptive
statistics of the types of features employed during consultations will
be reported, along with how much of an impact the lockdown has had on
different kinds of work and the changes made to work during the
pandemic. The open text questions exploring participant views towards
the use of technology when conducting consultations and how their work
has been impacted by the lockdown were thematically analysed. Inductive
thematic analysis was used to explore the semantic themes
[@boyatzisTransformingQualitativeInformation1998a]. Inductive analysis
is data driven and thus codes are derived from the data, rather than
pre-determined codes being used to analyse the data
[@guestAppliedThematicAnalysis2012]. The question assessing how much
each type of work has been affected will be plotted to check the
distribution of responses. Likert rating scales produce ordinal data and
the responses are unlikely to be normally distributed, therefore models
which assume normally distributed continuous data are inappropriate
[@liddellAnalyzingOrdinalData2017]. It will therefore be analysed using
a cumulative probit ordinal regression model
[@burknerOrdinalRegressionModels2019] to see if there differences
between types of work as well as groups, such as role. A probit model
will be used as the latent variable which the question is seeking to
measure (how greatly various kinds of work have been affected by the
lockdown) is assumed to be normally distributed
[@mccullaghRegressionModelsOrdinal1980].

To see whether the lockdown has affected the prevalence of different
features of consultation, a two-way within subjects ANOVA will be used.
This is suitable because the data is proportional
[@mangiaficoSummaryAnalysisExtension2016]. The features will be
clustered into subgroups: Solution-Focused, Problem analysis,
Organisation and knowledge; and Valuing everyone (see Appendix 3 for a
breakdown of the features into the subgroups). The subgroups will be
analysed independently with the other factor (time) common to all
subgroup analyses. Time will have two levels (before and during
lockdown). Responses which did not include any data for questions
relating to the features of consultation or how their work had been
impacted by the pandemic were excluded.

## 2.5 Reflections on pilot

### 2.5.1 Interview schedule

The interview was piloted with a Trainee Educational Psychologist (TEP)
to check for flow and whether the interviewees understood the questions.
The TEP commented on the definition of "features" in question 5. The
word was changed to "features" and a clarification statement will be
provided, along with a definition if necessary.

### 2.5.2 Observation schedule

To establish inter-rater reliability (IRR), an anonymous transcript of a
previously recorded consultation was analysed for feature using the
observation schedule. Three raters, including the researcher, assessed
the transcript for feature of consultations in their relative order.
Intraclass correlations [@shroutIntraclassCorrelationsUses1979] were
calculated between the three raters. The relative frequency of each
category was calculated for each rater and compared with one coder's
(the researchers) results. Because frequency counts were used,
intraclass correlations (ICC) were suitable as the data is continuous.
To calculate ICC, four factors must be decided upon prior to calculation
[@hallgrenComputingInterRaterReliability2012]. A two way model was used
because the raters weren't randomly selected from the population. Given
that a non-timed sequence record design was chosen for the observation
schedule, good IRR was defined as consistency in the ratings because it
was more important that raters provide scores that are similar in rank
order. A single measures ICC was calculated because the reliability of
the other two raters needed to generalise to ratings of one coder (the
researcher). And finally, a mixed model was used because the raters were
not randomly chosen from a population. This model was applied using the
irr package [@gamerIrrVariousCoefficients2019] in R. This produced an
ICC of 0.471 which, according to guidelines provided by
@cicchettiGuidelinesCriteriaRules1994 are 'substantial'.

The categories are defined and operationalised prior to data collection
[@crollSystematicClassroomObservation1986]. They were derived from the
relevant literature and were mutually exclusive. This was to increase
the reliability as it reduces the chances of observations being coded
differently according to the interpretation of an observer. To further
reduce risks to reliability, the categories were limited to what was
explicitly said. This was done to minimise the amount of inference the
researcher had to use when deciding whether a category was observed
[@crollSystematicClassroomObservation1986]. Models of consultation, such
as Solution-focused, were broken down into their constituent observable
parts, such as exploring strengths and identifying exceptions, so the
categories were more fine-grained and which specific features of the
models were used during consultations.

### 2.5.3 Questionnaire

14 TEPs piloted the questionnaire. They identified a few questions which
could be misinterpreted ("What will you do differently when things go
back to normal") and questions which would benefit from an explanation
as to how the answering mechanic worked. They also identified additional
types of work EPs could engage in during the lockdown. Overall the
feedback was positive, with particular focus on the breadth of questions
and the inclusion of changes to practice during the current lockdown.
One TEP reported that questions 17 and 18 could be interpreted in
different ways: do EPs think they should change their consultations
compared with how they were conducted prior to the lockdown or how they
are done now? Because the focus of the question is about what they will
do differently as a result of their experiences, the question was
changed to reflect this. A comment was raised about the order of the
options when selecting the magnitude of the impact on different kinds of
work due to the lockdown. It was decided it would be in ascending order
as this makes more intuitive sense when reading from left to right.

# 3 Results

NEED INTRODUCTION TO SECTION

```{=tex}
\newpage
\blandscape
```
![Thematic
map](D:/Doctorate/Assessments/Y3/Thesis/Results/Interview_transcripts/Figures/Thematic_map.png)

\elandscape

## 3.1 Interviews

30 EPs of varying roles and locations were interviewed. Participant's
roles included TEPs, maingrade EPs, specialist EPs, senior EPs, and
Principal EPs. The participants worked in locations such as London,
Yorkshire, Wales, and the Republic of Ireland. Thematic analysis
identified 32 inductive codes, as well as the 15 deductive codes
previously set, relating to what features EPs believed were effective
for consultation. 6 codes were identified for what made said features
effective (see Appendix XXX for the definitions of the inductive codes,
Appendix XXX for the definitions of the codes relating to what makes the
features effective and Appendix XXX for the breakdown of the number of
interviews which each code was identified in and the total number of
codes for each feature). These were combined to create 8 themes: Buy-in,
Conditions, Context, Strengths-based, Shared understanding,
Intervention, Future facing, and EP skills and knowledge. These could
then be combined to create two super themes: Internal factors (features relating to things happening with a consultation) and External factors (features relating to things happening around a consultation).

### 3.1.1 Buy-in

This theme related to the importance of EPs creating a bond with those
involved, including the consultee(s) and other school staff members not
directly involved in the consultation, and using this relationship to
facilitate change.

#### 3.1.1.1 Collaborative

One of the fundamental and most oft cited features for creating buy-in
was making consultation collaborative. Within the consultation, this was
achieved through a variety of factors. One of the key ones was making
sure there was equal participation, such that everyone had a voice and
different perspectives were heard: *"effective consultation shouldn't
being a meeting where one person dominates, whether that may be a
psychologist or anyone else"* (Interview 11) and *"it's like we're all
involved, we're all at the same level, we just come at it from a
different perspective"* (Interview 7).

As a result of there being equal participation, there is a greater
chance that everyone involved has the same understanding of the
situation and the CYP: *"to bring everyone together, and to co-create
and co-construct a shared narrative"* (Interview 11). Misunderstandings
can be cleared up (Interview 5) and these help everyone feel involved in
the process and ensure that the consultation is collaborative. The
creation of a shared narrative can also include the the creation of a
shared agenda. This helps guide the consultation so it is more effective
as it is meeting the needs of those involved and everyone agrees to it:
*"I think a really fundamentally important part of that consultation is
ensuring that we do have that shared agenda; we know why we're there
together and we all agree what we're doing there together"* (Interview
24) and *"to arrive at a joint action plan, joint for the school and the
parents, school are always involved as well, so it's more
collaborative"* (Interview 10).

This shared agenda can be established by identifying what everyone is
hoping to get from the consultation:

> *...it would always start with a question about what are your best hopes
> from our meeting together? What are your best hopes from our work
> together? Because if we don't start with that question, erm, then we
> don't know where we're trying to get to. (Interview 27)*

By working collaboratively with those involved, EPs can facilitate
collaboration between the home and school. This can potentially support
both by helping maintain morale and creating a sense of shared
responsibility:

> *...there is something that goes on often, not always, in the room when
> you've got the family, and school together, the, you do you do bring
> that sense of, 'We are working on this together; you are not alone
> school in this, you are not alone parents in this, we are doing this
> together'. (Interview 5)*

#### 3.1.1.2 Contributions valued

A related code, and one which can facilitate a collaborative
consultation, is the idea that everyone who is present in the
consultation should feel able to contribute. Not only this, but they
need to believe that what they say will be taken on board:

> *...where I would like to think that their views, their knowledge, their
> understanding is just as valid as mine... we are equal participants in
> this. (Interview 13)*

> *...equal participation, you know, as far as possible, or that everybody
> participates and that everybody feels valued, everybody feels that
> what they had to say is useful. (Interview 20)*

This can help give power to those who may not typically have it in
the school environment, thus helping create a more level playing field
and therefore a more collaborative consultation: *"schools are by
nature very hierarchical. So if you've got a TA they're often not seen
as the same as... a SENCO or a head teacher's views but in
that situation they are".* (Interview 1)

#### 3.1.1.3 Encouraging engagement

Removing power dynamics within a consultation was seen by many
participants as an important part of the EPs role within consultation.
This formed part of the code 'EP encouraging engagement'. The EP must
try and create a space so no consultee feels intimidated and in which
all relevant people can contribute, even if they cannot physically be
present:

> *...the psychologist trying to level power dynamics is a really key, a
> really key part of any consultation and that erm that's in relation to
> ourselves, as a professional with a doctorate normally, but also in
> relation to the family and the teacher, or the family and the school.
> (Interview 2)*

> *...balance of people's voices in the rooms. So, erm, making time for
> those that might not be able to be present in the meeting to hear
> their views and voices. (Interview 27)*

This code related to any effort by the EP to attempt to include the
voices of the relevant parties. One of the ways that this is through
*"active listening"* (Interview 1). A key idea related to the EP
facilitating others to participate:

> *I'm there to help facilitate the group in thinking about ways
> forward. (Interview 15)*

> *...giving a space where people can listen to other people's
> perspectives, then you take away the bulk of what it is that you're,
> erm, using to try and make a difference. (Interview 21)*

Not only does the EP need to facilitate others, but also challenge
potentially harmful narratives and navigate difficult situations:

> *Being careful and being prepared to challenge. (Interview 25)*

> *...sometimes a kind of mediation role because... we work in complex
> and messy situations. And it's not always that people are going to
> agree, or even really want to hear what they have to say. So there's
> that kind of control in the, the floor that happens in a consultation,
> which doesn't happen in other types of conversation. (Interview 3)*

Being able to read body language was identified by a few EPs as being
important for facilitating engagement:

> *you try to do an online meeting, you lose the gesticulations, you
> lose... being able to point at things or being able to... look at their faces better and realise, 'Oh, they're not
> understanding, I need to change the way I'm explaining it' or
> something. I think you lose so much because it's that non-verbal
> feedback that you get, that allows you to know where you are at with
> the relationship, to know the way you can develop within that
> consultation. (Interview 24)*

However, this was not universal. A few EPs found that using technologically-mediated (tech) consultations did not lead to a decrease in quality of the relationship. One EP experienced her consultees asking for telephone consultations and that these were effective (Interview 16).

#### 3.1.1.4 Rapport

The difference between in-person and tech consultations relates to
another core feature, which is the development of a rapport with those
involved in consultations. Within the consultation, an EP must quickly
develop a rapport so that the consultees feel comfortable talking about
potentially difficult topics:

> *...trust and credibility and shared mutual respect, I think are at the
> core of any consultation. You know, they value what I offer because
> I'm in touch and the fact they get on well with me, that almost
> therapeutic relationship. (Interview 7)*

> *...built up that trust and sense of safety, that it's okay to express
> their worries, that you can get quite a lot of information. (Interview
> 10)*

The EP needs to not only develop a rapport with those involved, but
encourage relationships between consultees: *"building attuned
interactions in a meeting with parents, with teachers, and then
hopefully between them as well. It just kind of gets everyone on the
same page, hopefully gets everyone pointing in the right direction"*
(Interview 30). This is especially important when relationships between
the home and school have broken down: *"sometimes you have a breakdown between parents and the school... you can be a person in between, and try and get that working through that... which is... a key feature of consultation. (Interview 4)*

Several EPs talked about the importance of having a good relationship
with the school. A good relationship between the school (generally
understood to mean at least the SECNCos and potentially Senior
Leadership Team) helps consultation to be more effective: *"If it's
going to be successful model in a school, I think the need is... time
for the EP to build a relationship with the school is important.*
(Interview 23). The reason the relationship is crucial for improving
consultation is that when the EP has developed a good relationship with
the school and they are mutually supporting one another, it is easier to
create an environment which fosters collaboration:

> *...when you know the school especially, and they're supporting you in
> supporting the parents and the staff to do that, then you see it a lot
> more". (Interview 1)*

> *...schools are often hesitant to adopt consultation as the main method
> of EP work: some of the SEN schools that I work with have a very rigid
> way of seeing the EP role and what we do, and they're, they're view
> is, more often than not, my role as an EP is to go in, do an
> assessment, write a report, and that's it. Er, so in those instances,
> I find it much harder to sell consultation as a, as a model.
> (Interview 11)*

However, several EPs spoke of using their relationship with the school
to change how they approach EP work and what the EP can do in the
school:

> *...once you build a relationship with schools, and you've been working
> in it, you can shift things, you can move things around, to, you know,
> working with a bit more control, getting them to see how, you know, it
> can be more effective, working with consultation, not doing just lots
> of assessments. (Interview 4)*

> *That's how you change it. I think that the relationship is super
> important. (Interview 23)*

#### 3.1.1.5 EP view of consultation and Consultee view of consultation

An important feature of consultation that relates to rapport is the
understanding that the consultees, EP, and school as a whole have
towards consultation. How the EP and consultees view consultation can
have a large impact on a consultation and its efficacy. A belief shared
by many interviewees was that *"both parties, kind of, know how
consultation works"* (Interview 24) and this *"might depend on people's
constructs of what consultation is"* (Interview 29). Interviewees had an
overwhelmingly positive view of consultation, highlighting its
versatility and alignment with their values:

> *...consultation, I think, is a, is a framework with the complexity that
> matches the complexity of the concerns that are being raised. Erm,
> we're looking at concerns at an individual and a group and a systemic
> level. (Interview 21)*

> *I don't think you can be inclusive without using a consultative
> model. (Interview 25)*

Though many interviewees identified the value of consultation and the
importance of clearly understanding it and what it involves, many also
pointed out that there is a large heterogeneity of practice among EPs:
*"I think that concept of what a consultation is will vary from one EP
to another"* (Interview 24). There are also EPs who do not value it and
prefer a more traditional style of assessing CYP and then writing a
report. As one interviewee said: *"I know there's a lot of EPs out there
that continue to work in that way and I think, I think that's one of the
barriers to shifting more to a consultation framework"* (Interview 17).
One interviewee, who had recently attended a course on consultation
provided by their EPS, stated:

> *I'm not sure a lot of EPs really understand what it is. Being able to
> communicate that... even on that consultation course that I mentioned
> I went on, I was really surprised that people, people very open and
> very honest, and they said, 'We've been saying we've been using
> consultation, but we actually have not. We've realised now that we
> haven't really been using consultation'. (Interview 22)*

This makes it difficult for consultees to gain a clear understanding of
what consultation is and has led a few EPs to call for clearer
communication and *"being better at communicating... what it is and what
it can do"* (Interview 22). One of the reasons it is important
consultees understand what consultation means is so they can see the
value in it. Many interviewees described how some of the schools they
work in do not appreciate it fully:

> *...if I could click my fingers and change something on a systemic level,
> it would be the attitude toward consultation because I I really view
> them as an investment. If you invest in a consultation, you're going
> to get better work and and outcomes. Whereas, sometimes they can be
> viewed as an expensive hurdle you have to get over to get a
> standardised score. (Interview 2)*

> *I think there are some schools that, erm, have a negative view of
> consultation. Because of that. It's, it's more complex procedure I
> think, people realise. (Interview 10)*

> *I think we need to educate our schools more about 'This is what the
> process is', because we say in sales blurb 'We do a consultation' and,
> erm, and then the schools are still stuck in that, kind of, old way of
> thinking. (Interview 28)*

A recurring comment centred around the differences between primary and
secondary schools, with primaries typically being more willing to engage
with them:

> *...most primary SENCOs are very open to whatever I suggest. And they're
> quite open to different ways of working, as long as they have a report
> to use as evidence, er, for EP involvement, so it has that element of
> of a tick box. But most primary schools are very open to different
> ways of looking, I would say, but secondaries definitely aren't.
> (Interview 18)*

#### 3.1.1.6 EP view of EPs and Consultee view of EPs

Another relevant strand to the different perceptions of consultations is
how the consultees view EPs and their role. Several interviewees talked
about how they were viewed as gatekeepers to resources or as someone who
would fix the situation independently of any work by the consultees:

> *...the associations that staff or parents can have of us as being, kind
> of, the deciders of resources. So we will go in and we will say, and
> we will think we are there to support to think about what we can do
> for this child, and they will think we are coming in to say 'Yes you
> can have any EHCP' or 'Yes you can have extra money'. (Interview 1)*

> *...if school are new to that way of working and they are used to having
> an EP come in and, sort of, tell them what to do. I do notice that
> sometimes there's a bit of confusion, er, especially from some
> teachers who are, 'Why are you asking me, aren't you supposed to tell
> me what I need to do'? (Interview 11)*

The consultee view of EPs also affected how receptive a school is to consultation because *"it very much comes down to the school's view of my role"* (Interview 14). Several EPs talked about wanting to change the views of the consultees in the consultation. ***

How the consultees view the EP can be changed in the consultation
itself: *"You're modelling how psychologists think... they might think a
psychologist is on a pedestal or whatever, but you're modelling that
psychologists are like everybody else"* (Interview 7). To help level
this power dynamic, EPs often try to present themselves as not having a
privileged position, as some interviewees talked about *"not putting
themselves in an expert position"* (Interview 27). This is because
*"It's the process of discussion itself, erm, that leads to, kind of,
outcomes, rather than taking on an expert model."* (Interview 14).
However, a few EPs pushed back against the framing of the EPs non-expert
stance as it can be counter-productive: *"I think, erm, sometimes EPs
can go too far the other way in not being the expert... it's a little
bit disingenuous, because sometimes we've got a lot of good ideas to
offer"* (Interview 27). How strongly they take on the role of the expert
was independent of the importance of most EPs placed on being empathetic
and supportive:

> *...you're in the situation as a human being, but also trying to be a
> psychologist as well, and they're quite difficult to do at the same
> time. (Interview 14)*

> *I think you need to be an ally, and a guide, but not be, 'I know what
> you should do and you should do this'. (Interview 23)*

#### 3.1.1.7 Willing to engage

A feature that almost two thirds of the interviewees identified was the
willingness of the consultees to engage in the consultation process:

> *...the effectiveness is because of engagement, critical thinking process
> thinking, and then plan your own action plans, which you're also
> engaged in. (Interview 5)*

> *...at the same time, to know that the reason that everyone is around the
> table for this consultation is to try and shift that thinking in some
> way. And usually, you know, just by nature of showing up everybody
> does want that, even if they don't necessarily believe it to be
> possible, which is why I think those features of consultation are
> effective. (Interview 3)*

> *...just general engagement from either the parents or school, and the
> willingness to, to change; the willingness to change their practice.
> (Interview 5)*

#### 3.1.1.8 Consultee ownership

Several interviewees talked about how these features are effective
because they help create a sense of consultee ownership of the
situation. By being collaborativeThe consultees are more likely to buy
into the process of consultation and are therefore more likely to feel
they can be an active agent in supporting the CYP:

> *...when people are active participants in a process, any process, they
> would be more likely to follow through with what has been agreed in
> terms of, whether that would be actions, whether that would be a
> specific approach that needs to be put in place. (Interview 11)*

> *...they retain some sense of ownership and some, er, sense of
> responsibility for putting in place what comes next. (Interview 20)*

> *...the point of that conversation is to leave something behind for the
> people who actually have power to do things and if you don't have
> their buy-in, then it's totally pointless. I'm struggling to think of
> a method, outside of consultation, where you could get that buy in and
> that information share and get to any kind of meaningful endpoint.
> (Interview 3)*

#### 3.1.1.9 Realistic

Another commonly discussed mechanism for effective consultations was the
increased chance of realistic recommendations and outcomes being
established. If the ideas generated are more co-constructed and built on
shared knowledge, they are more likely to be feasible:

> *...it also allows for reality, so if you've, you know, hopefully you're
> not getting ideas or strategies that are completely unworkable. So it
> should be based within the practice of the class teacher. So it isn't,
> you know, somebody coming in and going, 'Well, you need to do this
> three times a day with, you know, dah, dah, dah, dah, dah'. (Interview
> 21)*

> *...the feedback we get from parents that things are very grounded in
> reality, that the ideas that we're talking about makes sense because
> they come from a position of understanding and making sense of
> whatever is being brought into the room and, sort of, helping to
> manage some of the complexity. (Interview 27)*

#### 3.1.1.10 Consultees as experts

The final code from this theme relates to treating the consultees as
experts of their own area:

> *I try to make it collaborative because erm, my stance is that we all
> bring our own expertise; they're experts as parents, they're experts
> on their child. Erm and as teachers, they're experts on, you know,
> teaching that child and teaching in general. (Interview 8)*

> *I think they're effective because, we're capitalising on that idea
> that people are experts in their own lives. (Interview 22)*

### 3.1.2 EP skills and knowledge

The other most common theme related to the psychological knowledge and
skills EPs need to use when engaging in consultation.

#### 3.1.2.1 Knowledge

The most common code across all themes was in relation to the models of
consultation and general psychological knowledge that the interviewees
believed EPs needed to have to facilitate an effective consultation. The
*"use of theory and reference to the evidence base"* (Interview 2) was
identified as an important effective feature of consultation. Commonly
discussed models and frameworks included being solution-focused
(Interview 1), person-centred (Interview 16), trauma and attachment
informed (Interview 13), and using Wagner's model of consultation
(Interview 17) and the COMOIRA model (Interview 25). Other specific
psychological areas included using principles from Narrative Therapy
(Interview 17), an ecosystemic model (Interview 2), social
constructivism (Interview 6), as well as psychologies such as positive
psychology (Interview 9). Some interviewees saw their role as
*"sharing... and disseminating psychological theory"* (Interview 18) and
that consultation "helped [them] really use psychology with [their]
schools" (Interview 11).

The use of a model was often spoken positively as *"[giving] the
consultation a structure"* (Interview 11) and for one interviewee they
were the most important part:

> *...for me, the models of psychology are the number one priority, they
> have to be systemic and interactionist so that all behaviour is seen
> as a function of the person and the situation. So that if a concern is
> being described, we want to be looking at finding out about what was
> happening at the time or when it was happening. (Interview 27)*

#### 3.1.2.2 Presenting problem

Many EPs mentioned specific features within different models. One such
feature was exploring the presenting problem from the problem-analysis
framework [@monsenAccountableModelPractice1998]:

> *...getting an idea of what their main concerns are because when it feels
> very big, it's really the problem feels very big, the issue with the
> child is very messy. There's a lot going on, it can be hard to know
> where to start. So focusing them down is something that I do where I'm
> like 'What's your main concern?'. (Interview 8)*

This code also involved *"further clarification around the
difficulties"* (Interview 11) and a discussion of "What are the
conditions around it" (Interview 12).

#### 3.1.2.3 Treatments

Another code relating to the problem-analysis framework was the
discussion of treatments for the CYP. This involved *"planning
recommendations"* (Interview 2) and using the consultation *"as a space
where we can really drill down into exactly what you mean when you say
'A social skills group'"* (Interview 2) as you can decide what the
intervention is specifically for.

#### 3.1.2.4 Suggesting solutions

Another frequently mentioned model was the Solution-focused model
[@murphySolutionfocusedCounselingSchools1997]. A key part of this model
is suggesting solutions and several interviewees brought this idea up.
These are typically recommendations *"to be done at home and at school"*
(Interview 12) Several EPs stated they were happy to make
recommendations but simultaneously did not want to dominate the
consultation (Interview 11). The importance of taking on board what the
consultees said was also voiced by a few interviewee so that the EP does
not make recommendations that have already been tried (Interview 13).

#### 3.1.2.5 Exception seeking

Another code relating to the Solution-focused model was the discussion
of times when the main difficulty is reduced or absent:

> *...building all those principles of, yes, psychology that we're trained
> with, and we're taught to use: exception seeking (Interview 24)*

> *...finding out about other contexts when it was similar and other
> contexts when it was different, so that you're able to hypothesise
> about what's happening (Interview 27)*

#### 3.1.2.6 Reflective

A feature mentioned by almost all participants centred around the
importance of being reflective. This included the use of *"reflective
listening"* (Interview 1) and *"[checking] back in with people...
working with them just to understand, have they progressed on that
journey"* (Interview 16). Many interviewees brought up the importance of
checking with the consultees *"whether we did what we wanted to do, and
if not, what still needs to be done?"* (Interview 21).

The importance of being reflective was not limited to within the
consultation; the structure of consultation itself should also
incorporate reflection:

> *...it might be nice within models that we have with schools, if there's
> a definite agreement that there is follow up or a review, if it's not
> by me, if it's by someone in the school, because that, that, kind of,
> ensures that what's discussed in the consultation is actually, you
> know, implemented and monitored. (Interview 14)*

> *I also like to have a consultation as a feedback meeting at the end
> to... revisit what we've discussed in the first session,
> and obviously, by that time, I'll have gathered information from other
> sources to use that other information to further inform what is going
> to be done about the situation and to answer their referral question.
> (Interview 9)*

This reflective structure extends to the gaining of feedback from
consultees. The importance of feedback was mentioned by several
interviewees, for example: *"we have to treat it as a cyclical process
which has to be reviewed and evaluated so that we can use that feedback
to improve practices"* (Interview 1). Learning from peers through
observation and critical reflection with colleagues was also
highlighted:

> *I would hope that I'm a reflective practitioner and also, erm, having
> other people observe consultation, is really helpful in terms of
> trying to figure out, sometimes, what's going on, what made a
> difference. (Interview 21)*

> *...peer supervision is really helpful in terms of, er, helping your
> practice because obviously, you've got all that shared, sort of, ideas
> and knowledge and bouncing off each other in the team. (Interview 9)*

#### 3.1.2.7 Questioning

The use of question was discussed by more than half the interviewees,
using questions like *"'I wonder what would happen if?' 'What do you
think might happen if?'"* (Interview 25) to explore possibilities and
develop understanding. More banausic questions are asked to explore a
situation to gain a fuller understanding (Interview 5) as well as
exploring the context (Interview 27). However, as the consultation
progresses, questions can be used to get the consultees to think about
what change would look like for the CYP and how they could go about
achieving it (Interview 8). Not only is the content of the questioning
important, but the manner in which they are asked is a key factor *"how
EPs are asking those questions, and the types of questions they're
asking and, erm, the timing of those questions"* (Interview 15).

#### 3.1.2.8 Use of aids

A third of interviewees discussed types of supports that they use in
their consultations. Tools based on person-centred psychology, such as
Planning Alternative Tomorrows with Hope (PATHs) and Making Action Plans
(MAPs), was brought up by Interviewees 16 and 17. The use of metaphors
was endorsed as means to safely explore difficult topics (Interview 12),
as well as *"the Japanese problem-solving fish" and "blob trees"*
(Interview 10).

#### 3.1.2.9 Preparation

A number of interviewees highlighted the importance of being prepared
for a consultation, where the *"psychologist pools all that information,
formulates the hypothesis, types down what questions they want to ask"*
(Interview 20). This increases the chance of the consultation being
effective:

> *I think doing really thoughtful preparation is essential to to
> effective consultation, and I think sometimes there just isn't time
> for that but but really spending some time to think about, you know,
> what, what do we know? What what do I, what am I hoping to get out of
> this?" (Interview 13)*

This preparation, of *"being in the right headspace yourself"*
(Interview 13) extends to the consultees as them not being prepared can
hamper the efficacy of a consultation:

> *I think for a lot of the time, what limits my consultations is,
> they're just, sort of, caug- maybe a teacher is, sort of, caught on
> the cuff, they weren't really expecting me to meet them there, erm,
> but there they are. So they haven't really had time to, sort of,
> gather their thoughts beforehand. (Interview 20)*

#### 3.1.2.10 Reflexive

Another important feature of an effective consultation was the EP being
reflexive. This involves critically analysing, in the moment, *"'How's
my body language affecting the person that I'm speaking with?' 'How did
that question go down?' 'Was it understood?' Am I helping this
person?'"* (Interview 10). This process involves being *"flexible and
responsive"* (Interview 21) and *"adaptable"* (Interview 30) which can
be inhibited by the use of a consultation script (Interview 30). One
interviewee discussed the importance of:

> *...being aware of what's going on in the discussion and what the
> function of the discussion might be for the consultee at any one time.
> For example, if the adult is clearly struggling with the child, they
> might be looking for empathy... and understanding, so very much
> giving that but recognising that that in itself won't necessarily move
> things on. So trying to be aware... of what the function of their use
> of language is at the time and what they're trying to elicit from me.
> (Interview 14)*

Another interviewee discussed the importance of being sensitive to
*"anything that might be... difficult for potentially parents to talk
about"* (Interview 15). This process of being reflexive also helps
prevent the EP *"[imposing] my construct on the situation"* (Interview
25).

#### 3.1.2.11 Summarising

Several EPs mentioned that summarising or paraphrasing (Interview 19)
what has been said in a consultation is an important feature of
consultation. This includes *"re-speaking back to people what they've
told you"* (Interview 17) and *"[giving] a summary of what I think I've
heard from the different people"* (Interview 5).

#### 3.1.2.12 Efficient

The most frequently cited reason for consultation being effective was
that it is an efficient way of practising. This includes the simple fact
that *"more children get to have EP input"* (Interview 11) because it is
possible to *"talk about multiple students and put multiple things in
place as a result of that [consultation]"* (Interview 15). It is a tool
to *"gather information from different sources quickly"* (Interview 2)
which helps *"generate, hopefully accurate as possible, hypotheses"*
(Interview 30). Consultation also can *"effect change at a higher level
and a greater level"* (Interview 12) and there can be a *"ripple
effect... across policy level or across class or a group or even a whole
school"* (Interview 16). This means that "sometimes you might only need
one or *two consultation sessions to make some good change"* (Interview
17).

#### 3.1.2.13 Varied space for approach

Another key mechanism through which consultation is effective is its
versatility. *"Consultation is flexible"* (Interview 21) and a *"process
that evolves all the time"* (Interview 24). They allow for the use of
*"different strategies, different components"* (Interview 10) to meet
the needs of the consultees. Because consultation can be flexible, it
can adapt to a situation and therefore have a greater chance of a
positive impact:

> *I think the logistics of a consultation can remain the same, but the
> impact of a consultation can really vary. And... I don't
> know how many other tools we have available that that's the case for.
> So, if I think about the logistics for doing the BAS, or the logistics
> for doing a CBT session, I think that you very quickly become
> constrained by the way they were set up, whereas, the logistics for a
> consultation, getting some people in a room for a certain amount of
> time, allows for a flexibility. So... sometimes halfway
> through a consultation, you'll discover a piece of information that is
> crucial and up until now completely unknown, and you can change tack. (Interview 2)*

### 3.1.3 Shared understanding

This theme centres around the importance and ways in which EPs create a
common understanding of the situation between themselves and consultees.

#### 3.1.3.1 Different views

Almost every interviewee brought up the importance of gaining the views
of different people and *"gaining multiple perspectives"* (Interview
21). This includes *"the voice of the child, voice of the family, voice
of the teacher"* (Interview 17). It is particularly important to bring
the voice of the CYP: *"being quite child centred... bringing the
pupil voice into that discussion... [as] it's often not appropriate to
have the student in the room, especially if they're younger"* (Interview
15). A few interviewees talked about the importance of gaining the views
and including those with power in the system:

> *I think in some ways, as well, in consultation, making sure trying to
> involve, at some stage or at some level, people within a school or
> organisation who hold power. So that might be a head or a deputy head.
> Just because they have a lot of power within that system, to reframe.
> (Interview 17)*

> *...we are trying to become more active within the local authority as
> well. And I think that's very important. Because otherwise if you work
> as an EP service in isolation, without connects- strong connections
> with the senior leadership team within the local authority, and with
> the senior leadership team within the schools, nobody's gonna listen.
> (Interview 23)*

Consultation also allows for the *"understanding [of] different worlds
views, different cultural... constructs"* (Interview 17) and one
interviewee stated *"when people start to tell stories of things, it
gives you some quite good insights into how they think and where...
they're stuck in their thinking"* (Interview 12). By gaining different
views from consultees, the EP is better placed to make informed
hypotheses (Interview 20). When there is a disagreement between home and
school, consultation is an effective vehicle to *"bring that...
discrepancy into the room and discuss it and see if we can come up with
it with a kind of compromise or a way forward that... meets the needs of
both parties, and particularly for the student as well"* (Interview 15).

#### 3.1.3.2 Information gather

A related code was the EP gathering information not directly related to
the main concern: *"I find a lot out about the child, their background,
and... about the parents or family and what's going went around them"*
(Interview 8). This included *"[gathering] information from across the
four areas of SEND"* (Interview 2). This helps *"inform [their]
assessment"* (Interview 9). However, a number of interviewees made the
point that consultation is much more than simply gathering information:
*"the word 'consultation' might sometimes be interchangeably used with,
actually, what's really an information gathering process"* (Interview
24).

#### 3.1.3.3 Clarity

Over half the interviewees talked about the importance of gaining
clarity in a variety of ways. This included for *"what the process might
look like"* (Interview 20) as well as *"clarifying what people are
saying, what the parent is saying, what the SENCO is saying, what the
class teacher is saying"* (Interview 4). This done in the service of
*"understanding the situation better and exploring and understanding it
better"* (Interview 5). This allows for the EP to draw these strands
together and *"come to some kind of conceptualisation towards the end"*
(Interview 5).

#### 3.1.3.4 Setting out consultation plan

The establishing, by the EP, of the general structure for the
consultation was cited by more than half the interviewees as an
important feature of consultation. This was often done by exploring with
all those involved *"what we're hoping to get from the meeting, from the
consultation"* (Interview 24) because this *"gives it a clear
direction... [a] frame, [a] boundary"* (Interview 14). It also helps
*"[manages] everyone's expectations"* (Interview 14) and allows those
involved to know if they have achieved what they wanted to achieve
within the consultation (Interview 13).

#### 3.1.3.5 Language

Several interviewees brought up the importance of the language used
within a consultation. This had two main strands: potential language
difficulties due to English as a second language and the use of jargon
by the professional. One of the barriers to effective consultation is
*"lack of English language, from parents. It's not always possible to
have a translator... and even if you do... there are barriers... it's
difficult going through a third person. You have no idea... how
accurately they're translating"* (Interview 5). The other facet related
to the technical language that is pervades psychology and how this is
understood by the consultees:

*It takes a much higher level of skill to have a meaningful consultation
with somebody who does not have... the privilege of having... lots of
education, and... [a] big vocabulary and high level of verbal skill,
than it does... for us to sit around in a team surrounded by people who
are educated to doctorate level... But when you really need to try and
get meaningful information in a respectful way from from somebody who
finds language very hard, that's... a whole... nother level of
professional skill. (Interview 3)*

#### 3.1.3.6 Documentation

Documentation refers to the making of notes and summarising the contents
of the consultation. One EP stated it was *"[their] least favourite part
of the job... But unfortunately, it's really important, because I think
you've got an opportunity to write down and, kind of, what they call a
narrative, like rescue the words"* (Interview 17). Another expressed
more uniformly negative views towards documentation: *"what... will make
consultations: not having to flippin' write them up afterwards, we'd get
twice as many done... I don't understand why I'm writing about, the
magic happens in the room"* (Interview 22). However, others were more
positive: *"I think the written record is helpful of a consultation"*
(Interview 5) as it gives another opportunity to give advice at a later
date (Interview 30).

#### 3.1.3.7 EP explaining role

A small number of interviewees stated that making sure the consultees
understand what their role is within the consultation is important:
*"try and clarify what my role is and what it isn't"* (Interview 14).
This included *"[explaining] [their] involvement"* (Interview 14) and,
to help this process, one interviewee talked about *"[doing] role of the
EP insets, which we would offer every year, that talks about
consultation and the model of psychology and what's going to happen in
the meeting"* (Interview 27).

### 3.1.4 Intervention

Another theme which arose was the value of consultation as an intervention in and of itself. This was done through three mechanisms: providing a space for the EP to change consultees perceptions; emotionally supporting consultees, and consultation being part of the assessment process.

#### 3.1.4.1 Changing perspectives

One of the main ways in which interviewees talked about changing perspectives was around *"extending the thought processes of the people involved"* (Interview 10). A common idea among the interviewees was that the consultation *"facilitates that process of developing new meaning and new knowledge around a young person, or whatever the issue might be... reframing the way that people see it, which I think is a key element of change within consultation"* (Interview 17). The EP should also help others *"not [think] about a problem within a child, but [think] about a young person and how they interact with the environment that they are in"* (Interview 13). This change can also happen at a policy level, as one interviewee stated that consultation was the best vehicle to help schools become more inclusive (Interview 23). Consultation can also be used to help realign people's priorities and view towards those involved. Because of the highly pressurised nature of the systems we work in, *"family, and school can quite often fall out of sync and having a conversation together reminds everyone, they're on the same team"* (Interview 2).

This perspective change was not limited to the consultees views towards the CYP or situation; it extended to their views of consultation itself. One interviewee talked about how for *"[their] schools, once they were introduced to [consultation], and once they tried it, they really liked it"* and they could appreciate that *"consultation is a good model"* (Interview 11)

#### 3.1.4.2 Supporting consultees

Another point many EPs made was that consultations can often be used to help emotionally contain and provide support for the consultees: *...there is also something about consultation with schools that I find that can be emotionally containing for staff who perhaps are highly distressed (Interview 13)*. These *"therapeutic benefits"* (Interview 2) in a *"therapeutic style of meeting"* (Interview 22) often come through high levels of *"acceptance and empathy"* (Interview 13) because often consultees want to *"communicate with someone... how challenging it is for them"* (Interview 17). However, this was an area in which a few EPs judged that tech consultations were less effective as *"not being able to be physically there, as the sounding board, as their containing person... I couldn't be that... in a virtual environment"* (Interview 17)

#### 3.1.4.3 Assessment

A few interviewees saw the consultation as *"part of the assessment process"* (Interview 3) and as a *"powerful way to carry out assessment"* (Interview 19). This is because consultation can *"[lay] the foundation for an application for an EHCP assessment"* (Interview 2).

### 3.1.5 Strengths-based

Another emergent theme centred around the focus of consultation: it being strengths-based as it focuses on bringing out the skills of the consultees, highlighting what work is already having a positive impact for the cYP, and discussing the positive qualities of the CYP.

#### 3.1.5.1 Empowering individuals

One of the key features of an effective consultation is *"helping people to identify their own resources"* (Interview 10) and *"activate better existing skills and knowledge and competence"* (Interview 13)

#### 3.1.5.2 What's already working

One aspect which was frequently discussed was the exploration of what
was already working for the CYP. Interviewees talked about "[trying] to
build more of a strengths-based and positive outlook, and look at what's
working well, to shift things on" (Interview 22) and "trying to find
what has been tried, what has worked" (Interview 28).

#### 3.1.5.3 CYP strengths

The exploration of the strengths and positive qualities of the CYP was
also mentioned by several interviewees, such as: "it's exploring skills
and competencies alongside the problem" (Interview 27). A common idea was the consultations help reinvigorate the consultees and using the skills they already have:

> *"... building on what they potentially knew, but didn't really know what to do with it and... empowering and recognising that they were potentially able to sort out themselves". (Interview 19)*

> *"... a decent consultation... can help them feel empowered and perhaps a little bit reinfused about what their role could be". (Interview 13)*

A related idea was the empowering of those the consultees engage with, as a *"rising tide lifts all boats, in the sense that the person to whom I can give the consultation will very often generalise the advice from one case to another, from one session to another, from... one class to another"* (Interview 7)

### 3.1.6 Future facing

The final theme of the super code Internal factors focused on the idea of the consultation as helping to give a path forward for the consultees. This included the creation of goals for the CYP and the nature of consultation helping to prevent problems for other CYP in the future.

#### 3.1.6.1 Way forward

Over a third of the interviewees talked about how the nature of an effective consultation gives consultees a structure for how to move forward in supporting the CYP: *"it provides a mechanism to think about the future and to move forwards"* (Interview 15). Through consultation, the EP can *"elicit change or move people forward in a positive way"* (Interview 22) as well as identify the relevant support for the CYP (Interview 5). This is different from identifying specific goals for the CYP as consultations aren't *"always about solution finding because ways forwards aren't always solutions"* (Interview 3).

#### 3.1.6.2 Goal setting

For almost a third of interviewees, the identification of precise outcomes for the CYP to work towards is an important feature of consultation: 

> *"... for it to be consultation, I think there needs to be a clear, focus on finding, even if it's not a solution, but on coming up with a plan and... having a clear goal in mind". (Interview 11)*

> *"... [a] key component is goal setting, actually, and thinking about futures, and what the next steps would be". (Interview 17)*

However, one interviewee argued that not identifying clear goals does not *"necessarily make it an ineffective consultation"* (Interview 3).

#### 3.1.6.3 Preventative

Because of the emphasis on upskilling consultees within consultations, an EP using consultation can help prevent issues arising with other CYP within the school:

> *"when you're working with a teacher or with families or with different staff, actually the learning might be, the focus might be around a specific child, but actually that learning and that reframing can then be taken and be used preventatively with other young people or in the classroom" (Interview 17)*

By using consultations in different ways, such as regular features of school life, *"they would become more preventative"* (Interview 14).

#### 3.1.6.4 Ideas for future EP work

A few interviewees brought up the importance of using consultations to talk about and negotiate future EP involvement with the CYP (Interview 24). This might include an observation of the CYP in class (Interview 4).

### 3.1.7 Conditions

The first theme of the super code External factors related to the conditions of the consultations, including who was involved, how much time was set aside for the consultation, and the space in which it was held.

#### 3.1.7.1 Key people

Almost every interviewee cited having *"all the key stakeholders"* (Interview 11) involved in the consultation as a key aspect. Consultation was widely regarded as an *"indirect service method"* (Interview 17) so involved working with a range of people, including *"the SENCO, the class teacher, and both the mum and dad of that child"* (Interview 11). Many interviewees state that it was crucial to have *"the person that has most knowledge about the child"* (Interview 10) or the *"people who are most concerned"* (Interview 21). This included the person who *"has the most influence"* (Interview 14) as they will be the person who will implement the agreed interventions.

A number of interviewees highlighted the importance of bringing the voice of the CYP into the consultation, either by actively involving them in the consultation (Interview 21) or through those that know the CYP well (Interview 15).

Many interviewees identified difficulties with conducting consultations in secondary schools:

> *...if you've got multiple people working with a young person, and actually the more people you have, the less anybody feels any responsibility for them... you're trying to find that person who is most concerned and actually they don't exist. (Interview 13)*

> *...it's very difficult to get parents, teachers, parents and teachers around the same table, at the same time. (Interview 18)*

#### 3.1.7.2 Time

Over two thirds of the interviewees brought up time as an important part of a consultation. This mainly took the form of interviewees stating that the biggest barrier to effective consultations was a lack of time within the consultation, for example: *"I don't think you can have, say an, effective 20 minute consultation. It's not a consultation"* (Interview 26). This is because you need time for those involved to move beyond the "black and white way of thinking" about labels (Interview 18).

A related issue centred around the amount of time bought in by schools. Because the majority of interviewees either worked for fully traded services or as private EPs, the schools they worked with only had a limited amount of contact time. This led to several interviewees discussing the difficulty of bringing about change with schools because of the time limits placed on them (Interview 12). 

This was an area where tech consultations provided an advantage, as EPs can save time by not travelling between different schools (Interviews 13, 17, & 29).

#### 3.1.7.3 Resources

Resources was often cited important feature to consultations. This had several dimensions, including the ability of the consultees to enact change for the CYP due to resource constraints: 

> *The biggest barrier I come across is people saying, 'Well, that's lovely and I think we've come up with some fabulous ideas. However, I don't think management will let me do that'... So top down squashing... it's budgetary, it's time-bound, it's people saying, 'Well, we don't have the physical resources to be able to do that'. (Interview 16)*

> *... we might have all the ideas in the world around how someone might be supported. And it doesn't, I guess, affect the consultation in itself so much but it affects, it does affect the type of dialogue we might have around, schools and just the lack in, the workforce, the lack in staff, they're lacking the resourcing to really support some of these young people in the way that we would like them to be. I think that shapes consultations. (Interview 17)*

Another dimension is the resources school have to allow staff the time off from lessons to fully engage with a consultation: *"schools thinking 'We don't have the time and the capacity to free up staff to come and, come and sit and have a consultation'"* (Interview 15).

A third dimension related to the resources available to the schools to buy in EP time:

> *... I've certainly got schools that repeatedly say to me that they would love more EP time but they can't afford it in a traded environment and lots of... competing things that they have to spend money on. (Interview 21)*

#### 3.1.7.4 Space

A number of interviewees identified the importance of creating a space for effective consultations to occur. This encompassed both the physical space of the location and the mental space to be able to deal with complex experiences:

> *... sometimes people have asked to do consultations in rooms where there are other people and it's just messy. (Interview 2)*

>*I think the room that you meet in is quite important and the way that it's set up... so... it doesn't seem like an interview situation. (Interview 9)*

This aspect is particularly important for tech consultations as these almost always occur in the EP's and consultee's home:

> *... it can be difficult for staff to really, and parents, to really engage with the process, if they've got children running around and things going on. So... doing it where they can't have a separate space, emotionally as well as physically, can be tricky. (Interview 14)*

> *... having to make sure that doors are secure, so children can't run in at particular points. (Interview 24)*

#### 3.1.7.5 Confidential

Several interviewees brought up the importance of confidentiality for what was discussed in the consultation: *"we want to have a confidential place to reflect"* (Interview 22). This helps *"contribute towards building that kind of environment where people feel happy to share"* (Interview 15). The importance of confidentiality was made more important for many interviewees by the unexpected transition to tech consultations in response to the COVID-19 pandemic, with several identifying issues around security, for example: "the first step is finding an effective platform that's got enough safety features for us to be able to... carry out a consultation" (Interview 10).

### 3.1.8 Context

The second code within the External factors super theme was related to the general context that consultations are conducted within.

#### 3.1.8.1 Education systems

EPs work within many systems. These can all impact on individual consultations and on how EPs work through consultation. For example, *"there are schools who don't particularly value [consultation] and just want us to do assessments"* (Interview 5). As one interviewee stated, *"all the work of the EPs is determined by the context in which it's set and by the organisational agendas in which it's set"* (Interview 10). Several interviewees talked about the bureaucracy of the education system impacting on consultations and EP work as a whole:

> *...[the] role of the EP is less problem solving, it's more ticking a box, more bureaucratic exercise rather than a solving facilitation. (Interview 10)*

> *...there is a bureaucracy around an education, health care plan, in terms of certain reports being written, certain hurdles being gone through and certain assessments taking place. And so... we're not doing any thinking, we're merely following a bureaucratic process. (Interview 6)*

Other wider systemic issues related to how society as a whole sees additional needs:

> *...the medical model is so predominant... And I often find that those explanations for, learning and development and behaviours, can dominate conversations ADHD, ASD they are definitely a barrier to creating more effective, positive change. (Interview 17)*

> *...there's enormous pressure, ever increasing pressure on schools, to get results. And... [that's] antithetical to consultation. (Interview 25)*

A number of EPs identified operating in traded services as a barrier to consultation:

> *...I feel like it's the situation in which we work, the whole traded model, which means that consultation is, an addition... we just have to do it to get the information. It's not... valued as... a way of working in and of itself. (Interview 8)*

> *...I find within a traded service, you're quite constricted, in lots of ways about what the school expect in terms of the use of your time. (Interview 9)*

> *...I feel like... particularly in the traded service model, that dynamic is really hard to manage. And... it's been a real difficulty to introduce consultation as a working modelling in many of my schools. (Interview 11)*

Another issue that was identified was the views that school staff had towards change because of the people with more power in the system: *"SENCOs feeling unable to make change because of the head teacher"* (Interview 23).

#### 3.1.8.2 Individual differences

Almost five sixths of the interviewees brought up the characteristics of those involved in the consultation as an important feature of a consultation. The personalities, histories, and on the day mood of the consultees will likely impact on a consultation:

> *... that's going to play out in the room, in different ways, depending on the circumstances, the resilience of individuals, position, their own history, etc, etc. And will play out differently day to day, with the same people. (Interview 25)*

> *... there are parents who just don't like coming into school, are barred from school... have such a difficult relationship with school that is not possible. Physically can't get there because of health issues or younger children. (Interview 5)*

> *I think there's always going to be a level of... personality involved, that with some people, it is easier to... get that... feeling of engagement higher than it is with others... I think there is some variability, just because of human nature and the different personalities of the people that you meet. (Interview 12)*

The personality of the EP was identified by a few interviewees as potentially impacting on a consultation, for example: *"I think the personality of the individual EP can have a big impact"* (Interview 24). EP confidence in their own skill and knowledge was also identified as an important feature (Interview 14).

This variability in the presentation of consultation was viewed as a potential negative for consultation; if a teacher or parent was told they had to attend a consultation they *"wouldn't know what to expect because it would depend so much on the individual"* (Interview 11) because *"everybody has gone on their own and done totally different things"* (Interview 23).

#### 3.1.8.3 Understanding of SEN

A few interviewees stated that the way a school understands additional needs within an education context can have an impact on consultations. Some schools cleave to a more traditional 'within-child' understanding of additional needs, particularly secondary schools (Interview 14). As such, it is much harder to encourage these schools to adopt consultation as a way of working (Interview 11).

#### 3.1.8.4 EP workload

Almost a third of interviewees identified the amount of work EPs typically do as being a barrier to effective consultations. This was because the volume of work prohibits being able to fully engage with a case: 

>*... when you are on the day job, and you are 24-7 doing EP stuff, and you have... a stupid amount of cases and a stupid amount of schools and you cannot think... you are running on... empty (Interview 23)*

>*... you're so tired and stressed... you're not really thinking as well... you can't reflect on it and come up with different ideas and solutions because... you just have to get that written, get it sent off, and get on to the next thing. (Interview 9)* 

One interviewee identified the positive benefit of moving to tech consultations because *"I have a lot more time in my day, which means that I actually have a lot more space to think about children and cases"* (Interview 18). 

#### 3.1.8.5 School knowledge

A few interviewees stated that having *"in depth knowledge of schools and how they work"* (Interview 7), in particular secondary schools (Interview 22), helped their consultations be more effective. One interviewee explained that having good knowledge of the whole system differentiated EPs from clinical psychologists because EPs are "fluent... in that... understanding and situational context" (Interview 21).

#### 3.1.8.6 Outside system

Almost a third of interviewees stated that the EP working outside the school system helped their consultations be more effective: 

> *I think being an external person helps... you are able to ask some of the questions of parents that school can't, you can also ask questions of school that parents [can't] and take on that more challenging aspects. (Interview 5)*

> *... we have to get meta to the situation and not get too bogged down and immersed in the nitty gritty. So keeping meta and keeping perspective on it, I think is a skill that EPs can bring, that really helps. And that's the beauty of not working in the system, is the beauty of going in and out of schools. (Interview 27)*








POTENTIALLY USE ELSEWHERE The EP is gathering and summarising the ideas
and saying, 'Given what we've discussed, and the ideas we've heard so
far, what is going to make most sense for this young person and what's
going to make most difference'? And then it's getting the ideas from the
people. (Interview 27)

Collaboration was identified as a key factor not only because it
increased the consultee's willingness to engage but because it increased
the chances of the recommendations being put in place:

I think if you have a really good consultation and you can actually
problem solve together, and the people that you're consulting with,
actually come up with some of the ideas, then it's much more likely for
those interventions to happen. (Interview 20)

it underpins all of the work that we do with schools. So I would say
every school visit, team meetings, organisational level consultations,
we would be applying the same psychologies, the same frameworks.
(Interview 27)

#### 

## 3.2 Questionnaire

## Observations

No pair-wise simplifications could be made as there were no
consultations which saw change which differed by only 1 feature.

```{r Thematic map, echo = FALSE, eval = FALSE}
#This makes a graph of one theme but it doesn't look great (all the codes down below it) and I haven't given it any colour yet (need to work out how)
grViz("
digraph a_nice_graph {

# node definitions with substituted label text
node [fontname = Helvetica]
a [label = '@@1']
b [label = '@@2']
c [label = '@@3']
d [label = '@@4']
e [label = '@@5']

# edge definitions with the node IDs
a -> {b c d e}
}

[1]: 'Forward facing'
[2]: 'Goal setting'
[3]: 'Preventative'
[4]: 'Way forward'
[5]: 'Ideas for future EP work'
")

grViz("
digraph neato {

graph [layout = neato]

node [shape = circle,
      style = filled,
      color = grey,
      label = '']

node [fillcolor = green]
a

node [fillcolor = orange]
b c d e 

edge [color = grey]
a -> {b c d e}

#I can't just add the code from above which gives labels to the vector names from above.
#node [fontname = Helvetica]
#a [label = '@@1']
#b [label = '@@2']
#c [label = '@@3']
#d [label = '@@4']
#e [label = '@@5']
#
#[1]: 'Forward facing'
#[2]: 'Goal setting'
#[3]: 'Preventative'
#[4]: 'Way forward'
#[5]: 'Ideas for future EP work'

}")
```

# Discussion

Buy-in was facilitated by the EP not taking an expert stance and
creating a collaborative and sharing environment for the consultees to
explore their thoughts.

There was a large disparity between the number of inductive and deductive codes and the number of instances of each code, with the inductive codes being recorded more frequently than deductive codes. This suggests that the current literature does not accurately reflect how EPs are using consultation in practice.

# Appendices

## Appendix 1

1)  What is your role?
2)  How do you define consultation? What does it mean to you?
3)  What key words would you use?
4)  How often have you engaged with consultation?
5)  What history of consultation training do you have?
6)  Does your current EPS value consultation/operate a
    consultation-based service?
7)  Why do you use consultation?
8)  What do you believe are the key features of a consultation? What
    needs to be present for it to be more than a conversation?
9)  What features do you most frequently see (what is seen may be
    different what they believe is effective)?
10) What do you believe are the key features of an effective
    consultation (including examples)?
11) What makes them effective?
12) How could consultations be more effective?
13) What are the barriers to effective consultation?
14) If you could not use consultation, what work would you use instead?
15) What is the unique contribution of consultation?
16) What has changed with regards to your consultation work during
    lockdown?
17) How have you found this change?
18) Advantages/disadvantages?
19) Will you do anything differently after this is over?
20) Should the service/EPs as a whole do things differently?

## Appendix 2

+---------------------------------+------------------------------------+
| Category                        | Definition                         |
+=================================+====================================+
| Info gather                     | Fact finding or discussion of      |
|                                 | non-key concern(s).                |
+---------------------------------+------------------------------------+
| Suggesting solutions            | The EP volunteering a solution to  |
|                                 | the presenting concern.            |
+---------------------------------+------------------------------------+
| CYP strengths                   | Any discussion of the CYP's        |
|                                 | positive qualities: attributes,    |
|                                 | personality, actions, etc.         |
+---------------------------------+------------------------------------+
| Discussing what's already       | Discussion (including evaluation)  |
| working                         | of any intervention/change which   |
|                                 | has improved the current situation |
|                                 | for the CYP.                       |
+---------------------------------+------------------------------------+
| Everyone's contributions valued | Consultees giving their view on    |
|                                 | something e.g. presenting          |
|                                 | hypotheses, suggesting solutions,  |
|                                 | or the EP explicitly acknowledging |
|                                 | someone for their contribution.    |
|                                 | Not just the consultee(s)          |
|                                 | speaking/giving an answer to a     |
|                                 | factual question.                  |
+---------------------------------+------------------------------------+
| Understanding presenting        | Discussion of any aspect of the    |
| problem                         | main presenting concern(s)         |
|                                 | including scope, environmental     |
|                                 | factors, exceptions, etc. and why  |
|                                 | a problem may be present           |
|                                 | [@sheridanSchoolConsultation2000]  |
+---------------------------------+------------------------------------+
| Summarising                     | The EP saying back what has        |
|                                 | previously been stated by          |
|                                 | consultees in the consultation     |
|                                 | (potentially building on it but    |
|                                 | not necessarily).                  |
+---------------------------------+------------------------------------+
| Planning/implementing           | Discussion and agreement between   |
| treatments                      | the consultant and consultee on    |
|                                 | any interventions that will be     |
|                                 | implemented to support the CYP     |
|                                 | [@sheridanSchoolConsultation2000]. |
+---------------------------------+------------------------------------+
| EP using expert knowledge       | EP discussing topics which they    |
|                                 | have knowledge of (from both       |
|                                 | professional experience and        |
|                                 | academic reading) within school    |
|                                 | psychology theory and practice.    |
+---------------------------------+------------------------------------+
| EP explaining role              | EP explicitly talking about the    |
|                                 | work of an EP and its purpose.     |
+---------------------------------+------------------------------------+
| Setting out plan for            | Discussion of what will happen     |
| consultation                    | over the course of the             |
|                                 | consultation.                      |
+---------------------------------+------------------------------------+
| Ideas for future EP work        | Discussion of potential work an EP |
|                                 | can do in the future, such as      |
|                                 | consultation, assessment,          |
|                                 | observation, etc.                  |
+---------------------------------+------------------------------------+
| Empowering individuals          | Any comments or questions which    |
|                                 | aim to increase the skills of the  |
|                                 | consultees (teachers, parents,     |
|                                 | SENCOs, etc.)/upskilling           |
|                                 | consultees so they can solve their |
|                                 | problems                           |
|                                 | [@nolanProcessPsychologicalConsultation2014].|
|                                 |                                    |
+---------------------------------+------------------------------------+
| School knowledge                | Any comments or questions which    |
|                                 | increase understanding of how the  |
|                                 | school works.                      |
+---------------------------------+------------------------------------+

## Appendix 3

+------------------------+---------------------------------------------+
| Level                  | Feature                                     |
+========================+=============================================+
| Solution-focused       | Suggesting solutions; Highlighting the      |
|                        | strengths of the CYP; Discussing what is    |
|                        | already working; Exploring exceptions;      |
|                        | Suggesting ideas for future EP work.        |
+------------------------+---------------------------------------------+
| Problem analysis       | Fully understanding the presenting problem; |
|                        | How to implement the interventions          |
+------------------------+---------------------------------------------+
| Organisation and       | Gathering information; Summarising; Using   |
| knowledge              | knowledge; Setting out a plan; Explaining   |
|                        | what EPs do; School knowledge               |
+------------------------+---------------------------------------------+
| Valuing everyone       | Everyone contributing; empowering those     |
|                        | involved                                    |
+------------------------+---------------------------------------------+

## Appendix 4

+------------------------+---------------------------------------------+
| Code                   | Definition                                  |
+========================+=============================================+
| Assessment             | How consultation can be a form of           |
|                        | assessment.                                 |
+------------------------+---------------------------------------------+
| Changing perspectives  | Any discussion of the EP changing the       |
|                        | perspectives of consultees during           |
|                        | consultation or the understanding of        |
|                        | consultation by consultees.                 |
+------------------------+---------------------------------------------+
| Clarity                | Gaining clarity regarding the issues        |
|                        | through formulation etc.                    |
+------------------------+---------------------------------------------+
| Collaborative          | Any discussion of a joint or collaborative  |
|                        | aspect of consultation.                     |
+------------------------+---------------------------------------------+
| Confidential           | Confidentiality and privacy                 |
+------------------------+---------------------------------------------+
| Consultee view of      | How the consultees view consultation and    |
| consultation           | understand it, as well as discussion of     |
|                        | increasing understanding through training.  |
+------------------------+---------------------------------------------+
| Consultee views of EPs | How the consultee views the role of the EP, |
|                        | including as the expert.                    |
+------------------------+---------------------------------------------+
| Different views        | Gaining the views of a variety of different |
|                        | people, including the young person, to      |
|                        | explore narratives and triangulate          |
|                        | evidence.                                   |
+------------------------+---------------------------------------------+
| Documentation          | Writing of notes or reports which detail    |
|                        | what happened.                              |
+------------------------+---------------------------------------------+
| Education systems      | How the school systems and bureaucratic     |
|                        | processes of the British education system   |
|                        | impact consultation.                        |
+------------------------+---------------------------------------------+
| EP encouraging         | The EP being engaged in the consultation    |
| engagement             | through active listening to challenge       |
|                        | narratives and facilitate discussion.       |
+------------------------+---------------------------------------------+
| EP view of             | The EPs understanding of consultation.      |
| consultation           |                                             |
+------------------------+---------------------------------------------+
| EP view of EPs         | The EPs understanding of their role,        |
|                        | including as the expert.                    |
+------------------------+---------------------------------------------+
| EP workload            | How the high workload EPs experience        |
|                        | impacts consultation.                       |
+------------------------+---------------------------------------------+
| Goal setting           | Explicit discussion of outcomes and goal    |
|                        | setting.                                    |
+------------------------+---------------------------------------------+
| Individual differences | How the personalities and histories of the  |
|                        | consultees and consultors impacts           |
|                        | consultation.                               |
+------------------------+---------------------------------------------+
| Key people             | Having the people who are most concerned    |
|                        | present.                                    |
+------------------------+---------------------------------------------+
| Language               | Using language that can be understood by    |
|                        | all as well as issues regarding English as  |
|                        | an Additional Language.                     |
+------------------------+---------------------------------------------+
| Preparation            | Time for the consultees and consultors to   |
|                        | prepare.                                    |
+------------------------+---------------------------------------------+
| Preventative           | How consultation can help prevent issues    |
|                        | arising or exacerbating.                    |
+------------------------+---------------------------------------------+
| Questioning            | Use of a wide range of questions within     |
|                        | consultation for a multitude of purposes,   |
|                        | including to explore and challenge.         |
+------------------------+---------------------------------------------+
| Rapport                | The importance of relationships with those  |
|                        | involved and how it can be developed.       |
+------------------------+---------------------------------------------+
| Reflective             | Reflecting on an individual consultation,   |
|                        | receiving feedback, or having a review      |
|                        | consultation to explore how the situation   |
|                        | has progressed.                             |
+------------------------+---------------------------------------------+
| Reflexive              | In consultation checking, by the EP, of how |
|                        | they and others might be affected by the    |
|                        | discussion as well as what they are saying  |
|                        | and why.                                    |
+------------------------+---------------------------------------------+
| Resources              | How a lack of resources from the school can |
|                        | impact on consultation, including not       |
|                        | giving teachers enough time for them.       |
+------------------------+---------------------------------------------+
| Space                  | Having both the physical and mental space   |
|                        | to engage with consultation.                |
+------------------------+---------------------------------------------+
| Supporting consultees  | EPs providing therapeutic support for       |
|                        | consultees during a consultation.           |
+------------------------+---------------------------------------------+
| Time                   | Having enough time within the consultation  |
|                        | to maximise its use.                        |
+------------------------+---------------------------------------------+
| Understanding of SEN   | How consultees and schools see special      |
|                        | educational needs in children and how it    |
|                        | impacts consultation.                       |
+------------------------+---------------------------------------------+
| Use of aids            | Using aids such as Planning Alternative     |
|                        | Tomorrows with Hope etc.                    |
+------------------------+---------------------------------------------+
| Way forward            | General statements about how consultation   |
|                        | can provide a way forward.                  |
+------------------------+---------------------------------------------+
| Willing to engage      | Consultees being willing to engage with the |
|                        | process of consultation.                    |
+------------------------+---------------------------------------------+

## Appendix 5

+------------------------+---------------------------------------------+
| Code                   | Definition                                  |
+========================+=============================================+
| Consultee ownership    | Consultees having a sense of responsibility |
|                        | for what will happen next to support the    |
|                        | CYP.                                        |
+------------------------+---------------------------------------------+
| Consultees as experts  | Viewing consultees as experts in the lives  |
|                        | of the child or as teachers of the child    |
|                        | who have valuable knowledge to share.       |
+------------------------+---------------------------------------------+
| Efficient              | Being able to impact at multiple levels,    |
|                        | over time, and have wide ranging impacts.   |
+------------------------+---------------------------------------------+
| Outside system         | EPs being outside the school system giving  |
|                        | them a meta perspective, a new way of       |
|                        | seeing things, which allows them to         |
|                        | challenge and explore.                      |
+------------------------+---------------------------------------------+
| Realistic              | The recommendations made are realistic to   |
|                        | the setting and capabilities of those       |
|                        | involved, including regarding resources,    |
|                        | and are time bound.                         |
+------------------------+---------------------------------------------+
| Varied space for       | Consultation being a highly flexible        |
| approach               | vehicle to support CYP.                     |
+------------------------+---------------------------------------------+

## Appendix 6

| Code                              | File n | Total code n |
|-----------------------------------|--------|--------------|
| Everyone's contributions valued   | 14     | 33           |
| CYP strengths                     | 7      | 9            |
| Empowering individuals            | 19     | 68           |
| Exception seeking                 | 5      | 8            |
| EP explaining role                | 5      | 5            |
| Ideas for future EP work          | 4      | 4            |
| Information gathering             | 18     | 48           |
| EP using expert knowledge         | 30     | 223          |
| Understanding presenting problem  | 16     | 35           |
| School knowledge                  | 3      | 4            |
| Setting out plan for consultation | 16     | 31           |
| Suggesting solutions              | 11     | 14           |
| Summarising                       | 6      | 7            |
| Planning/ implementing treatments | 8      | 15           |
| Discussing what's already working | 11     | 21           |
| Assessment                        | 5      | 14           |
| Changing perspectives             | 25     | 118          |
| Clarity                           | 17     | 37           |
| Collaborative                     | 29     | 212          |
| Confidential                      | 10     | 13           |
| Consultee view of consultation    | 28     | 155          |
| Consultee views of EPs            | 26     | 84           |
| Different views                   | 27     | 150          |
| Documentation                     | 8      | 10           |
| Education systems                 | 27     | 134          |
| EP encouraging engagement         | 29     | 119          |
| EP view of consultation           | 22     | 77           |
| EP view of EPs                    | 14     | 28           |
| EP workload                       | 7      | 16           |
| Goal setting                      | 13     | 21           |
| Individual differences            | 24     | 47           |
| Key people                        | 27     | 81           |
| Language                          | 8      | 13           |
| Preparation                       | 10     | 22           |
| Preventative                      | 5      | 5            |
| Questioning                       | 19     | 43           |
| Rapport                           | 26     | 91           |
| Reflective                        | 26     | 110          |
| Reflexive                         | 9      | 21           |
| Resources                         | 15     | 22           |
| Space                             | 15     | 20           |
| Supporting consultees             | 12     | 27           |
| Time                              | 22     | 61           |
| Understanding of SEN              | 3      | 7            |
| Use of aids                       | 10     | 22           |
| Way forward                       | 13     | 22           |
| Willing to engage                 | 19     | 41           |
| Consultee ownership               | 15     | 27           |
| Consultees as experts             | 5      | 6            |
| Efficient                         | 18     | 43           |
| Outside system                    | 8      | 12           |
| Realistic                         | 7      | 11           |
| Varied space for approach         | 10     | 15           |

# References
